---
title: "Computer Lab 5: High Dimensional Data Analysis"
author: "Sara Kopelman"
date: today
date-format: long
format:
  pdf:
    pdf-engine: pdflatex
    toc: false
    number-sections: false
    booktabs: true
header-includes:
  - \usepackage{float}  
---

# Part 1: Eigenvalues and Eigenvectors

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(dplyr)
library(tidyr)
```

## 1)

We define a new variable $Y=E^\top X$, where $E=(e_1, e_2, e_3)$ is the matrix of eigenvectors of $\Sigma$. Then $Y_i=e_i^\top X$. We then compute the eigenvalues and the eigenvectors of the covariance matrix $\Sigma$.

```{r}
Sig <- cbind(c(1, -2, 0), c(-2, 5, 0), c(0,0,2))
eigen_results <- eigen(Sig)
```

## 2)

We want to verify the following relationships:

$$
\text{Var}(Y_i)=e_i^\top\Sigma e_i\\
\text{Cov}(Y_i,Y_k)=e_i^\top\Sigma e_k,\;i\neq k\\
\sigma_{11}+\dots+\sigma_{pp}=\sum_{i=1}^p\text{Var}(X_i)=\sum_{i=1}^p\text{Var}(Y_i)
$$

For the first relationship we have

$$
\text{Var}(Y_i)=\text{Var}(e_i^\top X)=e_i^\top\text{Var}(X)e_i=e_i^\top\Sigma e_i,
$$

for the second we have

$$
\text{Cov}(Y_i,Y_k)=\text{Cov}(e_i^\top X,e_k^\top X)=e_i^\top\text{Cov}(X,X)e_k=e_i^\top\text{Var}(X)e_k=e_i^\top\Sigma e_k,
$$

and for the final relationship we firstly first have

$$
\sigma_{11}+\dots+\sigma_{pp}=\sum_{i=1}^p\text{Var}(X_i)
$$

which is true by definition as the diagonal entries of $\Sigma$, $\sigma_{ii}$ correspond to $\text{Var}(X_i)$. We now need to verify that this equals $\sum_{i=1}^p\text{Var}(Y_i)$. Let $Y$ have the covariance matrix $\Lambda$. Then $$
\sum_{i=1}^p\text{Var}(Y_i)=\operatorname{tr}(\Lambda).
$$

Since $Y=E^\top X$ then

$$
\Lambda=\operatorname{Var}(E^\top X)=E^\top\Sigma E
$$

and then by the circular property of the trace

$$
\operatorname{tr}(\Lambda)=\operatorname{tr}(E^\top\Sigma E)=\operatorname{tr}(\Sigma EE^\top)=\operatorname{tr}(\Sigma)=\sum_{i=1}^p\operatorname{Var}(X_i)
$$

and we have thus verified all three relationships.

## 3)

We now want to verify that:

-   the first eigenvector has unit length

-   the first and second eigenvectors are orthogonal

We begin by extracting the eigenvectors:

```{r}
# Extract eigenvectors
E <- eigen_results$vectors

e1 <- E[, 1]
e2 <- E[, 2]
```

We verify the first statement by computing the length of the first unit vector by taking the norm $\Vert e_1\Vert=\sqrt{e_1^\top e_1}$ and we see that it indeed is equal to 1.

```{r}
#Check unit length of first eigenvector
norm_e1 <- sqrt(sum(e1^2))
norm_e1

```

We then check the orthogonality by verifying that $e_1^\top e_2=0$ which we also see is the case.

```{r}
#Check orthogonality between first and second eigenvectors
orthogonality <- sum(e1 * e2)
orthogonality
```

These properties are expected theoretically because the eigenvectors are orthonormal due to the covariance matrix being symmetric. Symmetric matrices admit an orthonormal eigenbasis.

# Part 2: Principal Component Analysis (PCA)

In this part we will perform principal component analysis on the weekly rates of returns for 5 stocks.

```{r}
data <- read.table("T8-4.DAT", header = FALSE)  # use header = FALSE if no header 
colnames(data) <- c("JP_Morgan", "Citibank", "Wells_Fargo", "Royal_Dutch_Shell", "ExxonMobil")


```

We begin by computing the principal components of the non-standardized data.

```{r}
PC <- prcomp (data, scale. = FALSE )
```

## 1)

We then determine the matrix of correlations between the principal components and the original variables, displayed in the below output.

```{r}
cor(data,PC$x)
```

We see that the first PC has a strong correlation between Royal Dutch Shell and Exxonmobil, while the second is mainly correlated with JP Morgan, Citibank and Wells Fargo. The other three PCs have smaller correlations overall but the 5th still has a moderately large correlation with Wells Fargo.

## 2-3)

We now want to analyze how many components capture most of the variability in the data and to evaluate and plot the proportion of variance explained. We do this first by plotting a scree plot, illustrated in @fig-scree, below. We observe that the first PC explains around 50% of the variance, while the second explains around 30%. The third explains just around 10%, while the last two explain less than that.

```{r}
#| echo: false
#| label: fig-scree
#| fig-cap: "Scree plot for principal components."
#| fig-pos: "H"
# Variance explained by each component
var_explained <- PC$sdev^2


# Proportion of variance explained
prop_var <- var_explained / sum(var_explained)

# Scree plot
plot(prop_var, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot")


```

Moving on to the proportion of variance explained, @fig-cum, below, displays this in the cumulative. We observe that the first two components alone explain 80% of the variance.

```{r}
#| echo: false
#| label: fig-cum
#| fig-cap: "Cumulative variance explained for principal components."
#| fig-pos: "H"
# Cumulative proportion of variance explained
cum_var <- cumsum(prop_var)
plot(cum_var, type = "b", xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     main = "Cumulative Variance Explained")
abline(h = 0.8, col = "red", lty = 2)  # 80% threshold
```

## 4)

We now move on to attempting to interpreting the principal components, focusing mainly on the first two. Reminding ourselves of the correlations between the PCs and the data from the previously displayed output we can surmise that PC1 represents the combined movement of the energy companies and could thus be regarded as the "energy-component". PC2 then represents variation among banks, and could be regarded as the "bank-component". The rest of the components are not as easy to interpret but we also saw that these first two alone manages to explain the majority of the variation alone.

## 5)

We now perform the same analysis again, but this time on standardized data. The below output displays the new correlation matrix between the PCs and the data. We observe that PC1 now reflects overall variation across all companies, with banks contributing slightly more than energy companies. Unlike the unstandardized PCA, no single sector dominates. On the other hand, PC2 now primarily captures the variation in the energy companies. The other PCs are more mixed, but PC3 is mainly correlated with JP Morgan and Wells Fargo. Specifically, PC3 shows large correlations of opposite signs for these banks, indicating that it captures a pattern of variation where they move in opposite directions relative to each other.

```{r}
# PCA on standardized data
PC_std <- prcomp(data, scale. = TRUE)


# For standardized data, loadings are already correlations
cor_matrix_std <- PC_std$rotation  # Each column ~ correlations with PCs
cor_matrix_std

# Proportion of variance explained
var_explained_std <- PC_std$sdev^2
prop_var_std <- var_explained_std / sum(var_explained_std)


```

We also once again plot a scree plot, and a cumulative variance plot, illustrated in Figure @fig-scree-std and @fig-cum-std, below. In the scree plot we observe that the first three PCs each capture about as much variance as before. We still explain about 80% of variance with the first two components, and about 90% with the first three.

```{r}
#| echo: false
#| label: fig-scree-std
#| fig-cap: "Scree plot for principal components from standardized data."
#| fig-pos: "H"

# Scree plot
plot(prop_var_std, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot (Standardized Data)")


```

```{r}
#| echo: false
#| label: fig-cum-std
#| fig-cap: "Cumulative variance explained for principal components from standardized data"
#| fig-pos: "H"
# Cumulative variance
cum_var_std <- cumsum(prop_var_std)
plot(cum_var_std, type = "b", xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     main = "Cumulative Variance Explained (Standardized Data)")
abline(h = 0.8, col = "red", lty = 2)

```

## 6)

In general, PCA is performed on standardized data to ensure that all variables contribute equally, preventing variables with large scales or high variance from dominating the first components. However, in the context of stocks, variance is a direct measure of volatility, which is inherently meaningful and may even be the focus of the analysis. Therefore, the choice between standardized and non-standardized PCA could depend on the goal: standardized PCA highlights patterns of co-movement across all stocks, while unstandardized PCA would include the risk contribution of the most volatile stocks.

# Part 3: Sparse Clustering

In this section we explore sparse clustering. We begin by simulating the data as detailed in the code below. The design is a two-cluster structure embedded in the first ten features, with the remaining sixty features consisting of random noise. Specifically, the first 25 observations are shifted by a constant in the first ten features, forming the first cluster, while the remaining 25 observations retain standard normal values, forming the second cluster.

```{r}
#| warning: false
library(sparcl)
set.seed(2025)
X <- matrix(rnorm(50*70), ncol=70)
X[1:25, 1:10] <- X[1:25, 1:10] + 1.5
X <- scale(X)
```

## 1)

We now turn to selecting the tuning parameter $s$, which controls the sparsity, for sparse K-means clustering. According to the explanations in the `sparcl` package documentation, the package provides a permutation-based approach utilizing the gap statistic (Tibshirani et al., 2001), which is used to determine the level of sparsity that best separates the clusters.

The algorithm for sparse K-means assigns a nonnegative weight $w_j$ to each feature and seeks clusters $C_1, \dots, C_K$ that maximize the weighted between-cluster sum of squares subject to sparsity constraints on the weight vector. For a given tuning parameter $s$, the objective function can be written as

$$
O(s) = \sum_{j=1}^p w_j \, BCSS_j,
$$

where $BCSS_j$ denotes the between-cluster sum of squares for feature $j$. The parameter $s$ controls sparsity through an $\ell_1$ constraint on the weights, $\Vert w\Vert_1\leq s$ with smaller values of $s$ leading to more zero weights and therefore a stronger feature selection.

The data is then randomly permuted $B$ times across each feature. This preserves the marginal distribution for each feature but destroys any dependence between features, which in turn deletes any potential cluster structures present in the data. For each permutation the objective function is maximized given the same parameter $s$ as for the original non-permuted data, giving us $O^*(s)$. The gap statistic is then defined as the log ratio between $O(s)$ and the average $O^*(s)$ across the $B$ permutations. I.e:

$$
\text{Gap}(s) = \log(O(s)) - \text{mean}(\log(O^*(s))).
$$

A large gap indicates that we have a strong cluster separation relative to random noise (which is what we have for the permuted data). A smaller gap indicates that any separation is similar to that resulting from random noise. The optimal tuning parameter $s$ is chosen then chosen as the value that maximizes $\text{Gap}(s)$, or alternatively as the smallest $s$ such that its gap statistic is within one standard deviation of the maximum:

$$
\text{Gap}(s) \ge \max_s \text{Gap}(s) - \text{sd}(\log(O^*(s))).
$$

In practice, this procedure is implemented in `sparcl` via the argument `wbounds`, which corresponds to the candidates of the tuning parameter that we consider. The code below executes this tuning-parameter selection procedure.

```{r}
set.seed(2025)
perm.out <- KMeansSparseCluster.permute(X, K = 2, nperms = 10, wbounds = seq(2, 7, by = 0.5))
# K = 2 clusters (you can adjust), nperms = number of permutations
# wbounds = range of tuning parameters to test

```

We observe that the best tuning parameter was determined to be 4.

```{r}
best_wbound <- perm.out$bestw
best_wbound

```

## 2)

We now perform the actual cluster analysis by the following:

```{r}
set.seed(2025)
sparse.out <- KMeansSparseCluster(X, K = 2, wbounds = best_wbound, nstart = 20)
# nstart = number of random starts for K-means

```

In order to evaluate the clustering procedure, we examine the feature weights, illustrated in @fig-cl-we below. As expected, the weights assigned to the first ten features are substantially larger than those for the remaining features, indicating that the clustering procedure correctly identified the features driving the cluster separation.

```{r}
#| echo: false
#| label: fig-cl-we
#| fig-cap: "Feature Weights from sparse cluster analysis with tuning parameter 4."
#| fig-pos: "H"
weights <- sparse.out[[1]]$ws

weights_df <- data.frame(
  Feature = seq_along(weights),
  Weight = weights
)

ggplot(weights_df, aes(x = Feature, y = Weight)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Sparse K-means Feature Weights",
    x = "Feature index",
    y = "Weight"
  )

```

We now repeat the procedure for a range of tuning parameter between 2 to 7.

```{r}
wbounds <- 2:7
weight_list <- list()

for (w in wbounds) {
  set.seed(2025)
  res <- KMeansSparseCluster(X, K = 2, wbounds = w, nstart = 20)
  weight_list[[as.character(w)]] <- res[[1]]$ws
}

```

@fig-cl-seq, below, illustrates the feature weights across tuning parameters, with the true signal features in red and the noise features in grey. For the smallest value of $s=2$, a few of the true signal features receive weights of zero due to the strong sparsity constraint. At $s=3$, all signal features receive nonzero weights while all noise features remain at zero, indicating that this tuning parameter fully separates out the relevant features. For $s=4$ and above, the weights remain constant and increasing $s$ further does not change the allocation. Interestingly, while the weights remain constant, they are not exactly equal among the signal features or among the noise features, even though their theoretical contribution to the cluster assignment is the same (or equally irrelevant for the noise features). Although the weight patterns suggest that $s=3$ would be the sparsest suitable choice, the permutation-based gap statistic selected $s=4$ as the optimal tuning parameter. At this level, a few noise features receive small nonzero weights. This occurs because the gap statistic evaluates the observed weighted BCSS relative to permuted data, and even small contributions from noise features can slightly increase this objective. In this way, the gap statistic balances cluster separation with sparsity, and the selected $s$ may not always correspond to the solution that is closest to the "real" values.

```{r}
#| echo: false
#| label: fig-cl-seq
#| fig-cap: "Sparse K-means feature weights across tuning parameters"
#| fig-pos: "H"
#| warning: false


# Convert weight_list to a data frame
weights_df <- do.call(rbind, weight_list) %>% as.data.frame()
weights_df$wbound <- wbounds
colnames(weights_df)[1:ncol(X)] <- paste0("Feature_", 1:ncol(X))

# Pivot to long format and add feature type
weights_long <- weights_df %>%
  pivot_longer(cols = starts_with("Feature_"), names_to = "Feature", values_to = "Weight") %>%
  mutate(
    Type = ifelse(as.numeric(gsub("Feature_", "", Feature)) <= 10, "Signal", "Noise")
  )

# Plot
ggplot(weights_long, aes(x = wbound, y = Weight, group = Feature, color = Type)) +
  geom_line(alpha = 0.8) +
  scale_color_manual(values = c("Signal" = "red", "Noise" = "grey70")) +
  theme_minimal() +
  labs(
    x = "Tuning Parameter (wbounds)",
    y = "Feature Weight",
    title = "Sparse K-means Feature Weights vs Tuning Parameter",
    color = "Feature Type"
  )

```

# Use of AI

AI (ChatGPT) has been utilized in this report primarily to assist with coding, including constructing and formatting plots, generating LaTeX code, and suggesting functions. Additionally, it provided light spelling and grammar checks, as well as suggestions for more clearly conveying results.
