---
title: "Computer Lab 4: High Dimensional Data Analysis"
author: "Sara Kopelman"
date: today
date-format: long
format:
  pdf:
    toc: false
    number-sections: false
    booktabs: true
header-includes:
  - \usepackage{float}  
documentclass: article
---

# Part 1: Eigenvalues and Eigenvectors

## 1)

We define a new variable $Y=E^\top X$, where $E=(e_1, e_2, e_3)$ is the matrix of eigenvectors of $\Sigma$. Then $Y_i=e_i^\top X$. We then compute the eigenvalues and the eigenvectors of the covariance matrix $\Sigma$

```{r}
Sig <- cbind(c(1, -2, 0), c(-2, 5, 0), c(0,0,2))
eigen_results <- eigen(Sig)
```

## 2)

We want to verify the following relationships:

$$
\text{Var}(Y_i)=e_i^\top\Sigma e_i\\
\text{Cov}(Y_i,Y_k)=e_i^\top\Sigma e_k,\;i\neq k\\
\sigma_{11}+\dots+\sigma_{pp}=\sum_{i=1}^p\text{Var}(X_i)=\sum_{i=1}^p\text{Var}(Y_i)
$$

For the first relationship we have

$$
\text{Var}(Y_i)=\text{Var}(e_i^\top X)=e_i^\top\text{Var}(X)e_i=e_i^\top\Sigma e_i
$$

and for the second we have

$$
\text{Cov}(Y_i,Y_k)=\text{Cov}(e_i^\top X,e_k^\top X)=e_i^\top\text{Cov}(X,X)e_k=e_i^\top\text{Var}(X)e_k=e_i^\top\Sigma e_k
$$

and for the final relationship first have

$$
\sigma_{11}+\dots+\sigma_{pp}=\sum_{i=1}^p\text{Var}(X_i)
$$

true by definition as the diagonal entries of $\Sigma$, $\sigma_{ii}$ correspond to $\text{Var}(X_i)$. We now need to verify that this equals $\sum_{i=1}^p\text{Var}(Y_i)$. Let $Y$ have the covariance matrix $\Lambda$. Then $$
\sum_{i=1}^p\text{Var}(Y_i)=\operatorname{tr}(\Lambda).
$$

Since $Y=E^\top X$ then

$$
\Lambda=\operatorname{Var}(E^\top X)=E^\top\Sigma E
$$

then by the circular property of the trace

$$
\operatorname{tr}(\Lambda)=\operatorname{tr}(E^\top\Sigma E)=\operatorname{tr}(\Sigma EE^\top)=\operatorname{tr}(\Sigma)=\sum_{i=1}^p\operatorname{Var}(X_i)
$$

and we have thus verified all three relationships.

## 3)

We now want to verify that:

-   the first eigenvector has unit length

-   the first and second eigenvectors are orthogonal

We begin by extracting the eigenvectors:

```{r}
# Extract eigenvectors
E <- eigen_results$vectors

e1 <- E[, 1]
e2 <- E[, 2]
```

We verify the first statement by computing the length of the first unit vector by taking the norm $\Vert e_1\Vert=\sqrt{e_1^\top e_1}$ and we see that it indeed is equal to 1.

```{r}
#Check unit length of first eigenvector
norm_e1 <- sqrt(sum(e1^2))
norm_e1

```

We then check the orthogonality by verifying that $e_1^\top e_2=0$ which we also see is the case.

```{r}
#Check orthogonality between first and second eigenvectors
orthogonality <- sum(e1 * e2)
orthogonality
```

These properties are expected theoretically because the eigenvectors are orthonormal due to the covariance matrix being symmetric. Symmetric matrices admit an orthonormal eigenbasis.

# Part 2: Principal Component Analysis (PCA)

In this part we will perform principal component analysis on the weekly rates of returns for 5 stocks.

```{r}
data <- read.table("T8-4.DAT", header = FALSE)  # use header = FALSE if no header 
colnames(data) <- c("JP_Morgan", "Citibank", "Wells_Fargo", "Royal_Dutch_Shell", "ExxonMobil")


```

We begin by computing the principal components of the non-standardized data.

```{r}
PC <- prcomp (data, scale. = FALSE )
```

## 1)

We then determine the matrix of correlations between the principal components and the original variables, displayed in the below output.

```{r}
# Standard deviations of original variables
sd_X <- apply(data, 2, sd)

# Matrix of correlations: cor(PCs, original variables)
cor_matrix <- sweep(PC$rotation, 2, PC$sdev, "*")  # multiply each column by PC sd
cor_matrix <- sweep(cor_matrix, 1, sd_X, "/")       # divide each row by original SD

cor_matrix
```

We see that the first PC has a strong correlation between Royal Dutch Shell and Exxonmobil, while the second is mainly correlated with JP Morgan, Citibank and Wells Fargo. The other three PCs have smaller correlations overall but the 5th still has a moderately large correlation with Wells Fargo.

## 2-3)

We now want to analyze how many components capture most of the variability in the data and to evaluate and plot the proportion of variance explained. We do this first by plotting a scree plot, illustrated in @fig-scree, below.

```{r}
#| echo: false
#| label: fig-scree
#| fig-cap: "Scree plot for principal components."
# Variance explained by each component
var_explained <- PC$sdev^2


# Proportion of variance explained
prop_var <- var_explained / sum(var_explained)

# Scree plot
plot(prop_var, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot")


```

```{r}
# Cumulative proportion of variance explained
cum_var <- cumsum(prop_var)
plot(cum_var, type = "b", xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     main = "Cumulative Variance Explained")
abline(h = 0.8, col = "red", lty = 2)  # 80% threshold
```

```{r}
# PCA on standardized data
PC_std <- prcomp(data, scale. = TRUE)

# 1. Matrix of correlations between PCs and original variables
# For standardized data, loadings are already correlations
cor_matrix_std <- PC_std$rotation  # Each column ~ correlations with PCs
cor_matrix_std

# 2. Proportion of variance explained
var_explained_std <- PC_std$sdev^2
prop_var_std <- var_explained_std / sum(var_explained_std)
prop_var_std

# 3. Plot proportion and cumulative variance explained
# Scree plot
plot(prop_var_std, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot (Standardized Data)")

# Cumulative variance
cum_var_std <- cumsum(prop_var_std)
plot(cum_var_std, type = "b", xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     main = "Cumulative Variance Explained (Standardized Data)")
abline(h = 0.8, col = "red", lty = 2)

```

# Part 3: Sparse Clustering

```{r}
library(sparcl)
set.seed(2025)
X <- matrix(rnorm(50*70), ncol=70)
X[1:25, 1:10] <- X[1:25, 1:10] + 1.5
X <- scale(X)
```

```{r}
set.seed(2025)
perm.out <- KMeansSparseCluster.permute(X, K = 2, nperms = 10, wbounds = seq(2, 7, by = 0.5))
# K = 2 clusters (you can adjust), nperms = number of permutations
# wbounds = range of tuning parameters to test

```

```{r}
best_wbound <- perm.out$bestw
best_wbound

```

```{r}
set.seed(2025)
sparse.out <- KMeansSparseCluster(X, K = 2, wbounds = best_wbound, nstart = 20)
# nstart = number of random starts for K-means

```

```{r}
wbounds <- 2:7
weight_list <- list()

for (w in wbounds) {
  set.seed(2025)
  res <- KMeansSparseCluster(X, K = 2, wbounds = w, nstart = 20)
  weight_list[[as.character(w)]] <- res[[1]]$ws
}

```

```{r}
library(ggplot2)
library(reshape2)

# wbounds = 2:7
wbounds <- 2:7
weight_list <- list()

for (w in wbounds) {
  set.seed(2025)
  res <- KMeansSparseCluster(X, K = 2, wbounds = w, nstart = 20)
  weight_list[[as.character(w)]] <- res[[1]]$ws
}

# Convert weight_list to a data frame
weights_df <- do.call(rbind, weight_list)
weights_df <- as.data.frame(weights_df)
weights_df$wbound <- wbounds  # explicitly add the tuning parameter column

colnames(weights_df)[1:ncol(X)] <- paste0("Feature_", 1:ncol(X))

# Melt for ggplot
weights_melt <- melt(weights_df, id.vars = "wbound", variable.name = "Feature", value.name = "Weight")

# Plot
ggplot(weights_melt, aes(x = wbound, y = Weight, color = Feature)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Tuning Parameter (wbound)", y = "Feature Weight",
       title = "Sparse K-means Feature Weights vs Tuning Parameter") +
  theme(legend.position = "none")  # Hide legend if too many features

```
