---
title: "Computer Lab 5: High Dimensional Data Analysis"
author: "Sara Kopelman"
date: today
date-format: long
format:
  pdf:
    pdf-engine: pdflatex
    toc: false
    number-sections: false
    booktabs: true
header-includes:
  - \usepackage{float}  
---

# Part 1: Eigenvalues and Eigenvectors

## 1)

We define a new variable $Y=E^\top X$, where $E=(e_1, e_2, e_3)$ is the matrix of eigenvectors of $\Sigma$. Then $Y_i=e_i^\top X$. We then compute the eigenvalues and the eigenvectors of the covariance matrix $\Sigma$

```{r}
Sig <- cbind(c(1, -2, 0), c(-2, 5, 0), c(0,0,2))
eigen_results <- eigen(Sig)
```

## 2)

We want to verify the following relationships:

$$
\text{Var}(Y_i)=e_i^\top\Sigma e_i\\
\text{Cov}(Y_i,Y_k)=e_i^\top\Sigma e_k,\;i\neq k\\
\sigma_{11}+\dots+\sigma_{pp}=\sum_{i=1}^p\text{Var}(X_i)=\sum_{i=1}^p\text{Var}(Y_i)
$$

For the first relationship we have

$$
\text{Var}(Y_i)=\text{Var}(e_i^\top X)=e_i^\top\text{Var}(X)e_i=e_i^\top\Sigma e_i
$$

and for the second we have

$$
\text{Cov}(Y_i,Y_k)=\text{Cov}(e_i^\top X,e_k^\top X)=e_i^\top\text{Cov}(X,X)e_k=e_i^\top\text{Var}(X)e_k=e_i^\top\Sigma e_k
$$

and for the final relationship first have

$$
\sigma_{11}+\dots+\sigma_{pp}=\sum_{i=1}^p\text{Var}(X_i)
$$

true by definition as the diagonal entries of $\Sigma$, $\sigma_{ii}$ correspond to $\text{Var}(X_i)$. We now need to verify that this equals $\sum_{i=1}^p\text{Var}(Y_i)$. Let $Y$ have the covariance matrix $\Lambda$. Then $$
\sum_{i=1}^p\text{Var}(Y_i)=\operatorname{tr}(\Lambda).
$$

Since $Y=E^\top X$ then

$$
\Lambda=\operatorname{Var}(E^\top X)=E^\top\Sigma E
$$

then by the circular property of the trace

$$
\operatorname{tr}(\Lambda)=\operatorname{tr}(E^\top\Sigma E)=\operatorname{tr}(\Sigma EE^\top)=\operatorname{tr}(\Sigma)=\sum_{i=1}^p\operatorname{Var}(X_i)
$$

and we have thus verified all three relationships.

## 3)

We now want to verify that:

-   the first eigenvector has unit length

-   the first and second eigenvectors are orthogonal

We begin by extracting the eigenvectors:

```{r}
# Extract eigenvectors
E <- eigen_results$vectors

e1 <- E[, 1]
e2 <- E[, 2]
```

We verify the first statement by computing the length of the first unit vector by taking the norm $\Vert e_1\Vert=\sqrt{e_1^\top e_1}$ and we see that it indeed is equal to 1.

```{r}
#Check unit length of first eigenvector
norm_e1 <- sqrt(sum(e1^2))
norm_e1

```

We then check the orthogonality by verifying that $e_1^\top e_2=0$ which we also see is the case.

```{r}
#Check orthogonality between first and second eigenvectors
orthogonality <- sum(e1 * e2)
orthogonality
```

These properties are expected theoretically because the eigenvectors are orthonormal due to the covariance matrix being symmetric. Symmetric matrices admit an orthonormal eigenbasis.

# Part 2: Principal Component Analysis (PCA)

In this part we will perform principal component analysis on the weekly rates of returns for 5 stocks.

```{r}
data <- read.table("T8-4.DAT", header = FALSE)  # use header = FALSE if no header 
colnames(data) <- c("JP_Morgan", "Citibank", "Wells_Fargo", "Royal_Dutch_Shell", "ExxonMobil")


```

We begin by computing the principal components of the non-standardized data.

```{r}
PC <- prcomp (data, scale. = FALSE )
```

## 1)

We then determine the matrix of correlations between the principal components and the original variables, displayed in the below output.

```{r}
# Standard deviations of original variables
sd_X <- apply(data, 2, sd)

# Matrix of correlations: cor(PCs, original variables)
cor_matrix <- sweep(PC$rotation, 2, PC$sdev, "*")  # multiply each column by PC sd
cor_matrix <- sweep(cor_matrix, 1, sd_X, "/")       # divide each row by original SD

cor_matrix
```

We see that the first PC has a strong correlation between Royal Dutch Shell and Exxonmobil, while the second is mainly correlated with JP Morgan, Citibank and Wells Fargo. The other three PCs have smaller correlations overall but the 5th still has a moderately large correlation with Wells Fargo.

## 2-3)

We now want to analyze how many components capture most of the variability in the data and to evaluate and plot the proportion of variance explained. We do this first by plotting a scree plot, illustrated in @fig-scree, below. We observe that the first PC explains around 50% of the variance, while the second explains around 30%. The third explains just around 10%, while the last two explain less than that.

```{r}
#| echo: false
#| label: fig-scree
#| fig-cap: "Scree plot for principal components."
# Variance explained by each component
var_explained <- PC$sdev^2


# Proportion of variance explained
prop_var <- var_explained / sum(var_explained)

# Scree plot
plot(prop_var, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot")


```

Moving on to the proportion of variance explained, @fig-cum, below, displays this in the cumulative. We observe that the first two components alone explain 80% of the variance.

```{r}
#| echo: false
#| label: fig-cum
#| fig-cap: "Cumulative variance explained for principal components."
# Cumulative proportion of variance explained
cum_var <- cumsum(prop_var)
plot(cum_var, type = "b", xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     main = "Cumulative Variance Explained")
abline(h = 0.8, col = "red", lty = 2)  # 80% threshold
```

## 4)

We now move on to attempting to interpreting the principal components, focusing mainly on the first two. Reminding ourselves of the correlations between the PCs and the data from the previously displayed output we can surmise that PC1 represents the combined movement of the energy companies and could thus be regarded as the "energy-component". PC2 then represents variation among banks, and could be regarded as the "bank-component". The rest of the components are not as easy to interpret but we also saw that these first two alone manages to explain the majority of the variation alone.

## 5)

We now perform the same analysis again, but this time on standardized data. The below output displays the new correlation matrix between the PCs and the data. We observe that PC1 now reflects overall variation across all companies, with banks contributing slightly more than energy companies. Unlike the unstandardized PCA, no single sector dominates. On the other hand, now primarily captures the variation in the energy companies. The other PCs are more mixed, but PC3 is mainly correlated with JP Morgan and Wells Fargo. Specifically, PC3 shows large correlations of opposite signs for these banks, indicating that it captures a pattern of variation where they move in opposite directions relative to each other.

```{r}
# PCA on standardized data
PC_std <- prcomp(data, scale. = TRUE)


# For standardized data, loadings are already correlations
cor_matrix_std <- PC_std$rotation  # Each column ~ correlations with PCs
cor_matrix_std

# Proportion of variance explained
var_explained_std <- PC_std$sdev^2
prop_var_std <- var_explained_std / sum(var_explained_std)


```

We also once again plot a scree plot, and a cumulative variance plot, illustrated in Figure @fig-scree-std and @fig-cum-std, below. In the scree plot we observe that the first three PCs each capture about as much variance as before. We still explain about 80% of variance with the first two components, and about 90% with the first three.

```{r}
#| echo: false
#| label: fig-scree-std
#| fig-cap: "Scree plot for principal components from standardized data."

# Scree plot
plot(prop_var_std, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot (Standardized Data)")


```

```{r}
#| echo: false
#| label: fig-cum-std
#| fig-cap: "Cumulative variance explained for principal components from standardized data"
# Cumulative variance
cum_var_std <- cumsum(prop_var_std)
plot(cum_var_std, type = "b", xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     main = "Cumulative Variance Explained (Standardized Data)")
abline(h = 0.8, col = "red", lty = 2)

```

## 6)

In general, PCA is performed on standardized data to ensure that all variables contribute equally, preventing variables with large scales or high variance from dominating the first components. However, in the context of stocks, variance is a direct measure of volatility, which is inherently meaningful and may even be the focus of the analysis. Therefore, the choice between standardized and non-standardized PCA could depend on the goal: standardized PCA highlights patterns of co-movement across all stocks, while unstandardized PCA would include the risk contribution of the most volatile stocks.

# Part 3: Sparse Clustering

In this section we explore sparse clustering. We begin by simulating the data as detailed in the below output.

```{r}
library(sparcl)
set.seed(2025)
X <- matrix(rnorm(50*70), ncol=70)
X[1:25, 1:10] <- X[1:25, 1:10] + 1.5
X <- scale(X)
```

## 1)

We now turn to selecting the tuning parameter for sparse K-means clustering. Following the explanations in the `sparcl` package documentation, the package provides a permutation-based approach inspired by the gap statistic (Tibshirani et al., 2001), which helps identify the level of sparsity that best separates the clusters.

Sparse K-means assigns a weight $w_j$ to each feature and seeks clusters $C_1, \dots, C_K$ that maximize the weighted between-cluster sum of squares. Formally, the objective is

$$
O(s) = \sum_{j=1}^p w_j \, BCSS_j,
$$

where $BCSS_j$ measures the contribution of feature $j$ to cluster separation, and $s$ controls the overall sparsity of the weights via an L1 constraint.

To evaluate whether the observed clustering is stronger than expected by chance, the method permutes the values within each feature and recomputes the objective, yielding a set of permuted values $O^*(s)$. The gap statistic is then defined as

$$
\text{Gap}(s) = \log(O(s)) - \text{mean}(\log(O^*(s))).
$$

A larger gap indicates that the clustering is more distinct relative to random noise. The optimal tuning parameter $s$ can be chosen as the one that maximizes $\text{Gap}(s)$, or alternatively as the smallest $s$ for which the gap is within one standard deviation of the maximum:

$$
\text{Gap}(s) \ge \max_s \text{Gap}(s) - \text{sd}(\log(O^*(s))).
$$

The below code executes the procedure.

```{r}
set.seed(2025)
perm.out <- KMeansSparseCluster.permute(X, K = 2, nperms = 10, wbounds = seq(2, 7, by = 0.5))
# K = 2 clusters (you can adjust), nperms = number of permutations
# wbounds = range of tuning parameters to test

```

We see that the best *wbound* was determined to be 4.

```{r}
best_wbound <- perm.out$bestw
best_wbound

```

## 2)

We now perform the actual cluster analysis

```{r}
set.seed(2025)
sparse.out <- KMeansSparseCluster(X, K = 2, wbounds = best_wbound, nstart = 20)
# nstart = number of random starts for K-means

```

```{r}
wbounds <- 2:7
weight_list <- list()

for (w in wbounds) {
  set.seed(2025)
  res <- KMeansSparseCluster(X, K = 2, wbounds = w, nstart = 20)
  weight_list[[as.character(w)]] <- res[[1]]$ws
}

```

```{r}
library(ggplot2)
library(reshape2)

# wbounds = 2:7
wbounds <- 2:7
weight_list <- list()

for (w in wbounds) {
  set.seed(2025)
  res <- KMeansSparseCluster(X, K = 2, wbounds = w, nstart = 20)
  weight_list[[as.character(w)]] <- res[[1]]$ws
}

# Convert weight_list to a data frame
weights_df <- do.call(rbind, weight_list)
weights_df <- as.data.frame(weights_df)
weights_df$wbound <- wbounds  # explicitly add the tuning parameter column

colnames(weights_df)[1:ncol(X)] <- paste0("Feature_", 1:ncol(X))

# Melt for ggplot
weights_melt <- melt(weights_df, id.vars = "wbound", variable.name = "Feature", value.name = "Weight")

# Plot
ggplot(weights_melt, aes(x = wbound, y = Weight, color = Feature)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Tuning Parameter (wbound)", y = "Feature Weight",
       title = "Sparse K-means Feature Weights vs Tuning Parameter") +
  theme(legend.position = "none")  # Hide legend if too many features

```
