---
title: "Computer Lab 3: High Dimensional Data Analysis"
author: "Sara Kopelman"
date: today
date-format: long
format:
  pdf:
    toc: false
    number-sections: false
    booktabs: true
header-includes:
  - \usepackage{float}  
---

# Part 1: Multivariate regression analysis

## 1)

We begin by analyzing the amitriptyline overdose data. @fig-scatter, below, illustrates scatter plots between all the variables in the data-set. We observe that some input variables, such as $Z_2$ and $Z_3$, show indications of being correlated with each other. We also see that the two outputs, $Y_1$ and $Y_2$, seem to be clearly correlated, implying that a multivariate multiple regression approach may be worthwhile.

```{r}
#| echo: false
#| fig-cap: "Scatter plots of the variables in the amitriptyline overdose dataset"
#| label: fig-scatter
#| fig-pos: "H"
#| results: 'hide'
library(knitr)
library(kableExtra)
data.df <- read.table("ami.data.txt")
names <- c("Y1","Y2","Z1","Z2","Z3","Z4","Z5")
colnames(data.df) <- names
summary(data.df)
pairs(data.df)
```

## 2)

To begin with, we perform a univariate regression of $Y_1$ on $Z_1$–$Z_5$. @tbl-mod-1 and [-@tbl-mod1-metrics], below, display the estimated coefficients and diagnostics for this model. We observe that all coefficients, except for $\hat\beta_5$, are statistically different from 0 at a significance level of $\alpha = 0.05$. The R-squared is also relatively high at 0.88.

```{r}
#| echo: false
#| label: tbl-mod-1
#| tbl-cap: "Regression coefficients for model Y1"
model.Y1 <- lm(Y1 ~ Z1 + Z2 + Z3 + Z4 + Z5, data = data.df)
s <- summary(model.Y1)

# Extract coefficients
coef_table <- s$coefficients
kable(coef_table)

```

```{r}
#| echo: false
#| label: tbl-mod1-metrics
#| tbl-cap: "Regression coefficients for model Y1"
# Optional: show other summary metrics
summary_metrics <- data.frame(
  Metric = c("Residual Std. Error", "R-squared", "Adjusted R-squared", "F-statistic"),
  Value = c(s$sigma, s$r.squared, s$adj.r.squared, s$fstatistic[1])
)
kable(summary_metrics)


```

@fig-mod1-diag, below, illustrates residual diagnostic plots for the model. The Q-Q plot indicates that the residuals are approximately normally distributed. The Residuals vs Leverage plot also looks acceptable, although there are a few influential observations that may be worth investigating further. However, the Residuals vs Fitted and Scale-Location plots show indications of heteroscedasticity in the residuals.

```{r}
#| echo: false
#| label: fig-mod1-diag
#| fig-cap: "Diagnostic plots for model Y1"
#| fig-pos: "H"
#| fig-width: 8
#| fig-height: 8
# Base R diagnostic plots
par(mfrow = c(2, 2))  # Arrange 4 plots in 2x2 grid
plot(model.Y1)
par(mfrow = c(1,1))   # Reset to single plot
```

We also constructed a prediction interval for a new observation $z = (1, 1250, 150, 70, 85)^\top$. The resulting prediction and the interval boundaries are shown in @tbl-mod1-predint, below.

```{r}
#| echo: false
#| label: tbl-mod1-predint
#| tbl-cap: "95% Prediction Interval for new observation"
new.df        <- data.frame(Z1 =1, Z2 = 1250, Z3 = 150, Z4 = 70, Z5 = 85)
prediction.Y1 <- predict(model.Y1, new.df,interval = "prediction", level = 0.95)
kable(prediction.Y1)
```

## 3)

Having fitted a single multiple regression model, we next jointly regress $Y_1$ and $Y_2$ on the covariates. The resulting coefficients are displayed in @tbl-mod-coef, below. Note that the estimates for $Y_1$ are the same as in the univariate multiple regression model. We also observe that the number of significant coefficients in the model for $Y_2$ is smaller compared to $Y_1$, but the estimates themselves are quite similar for all variables except $Z_5$.

```{r}
#| echo: false
#| label: tbl-mod-coef
#| tbl-cap: "Regression coefficients for model Y1 and Y2"
#| warning: false
#| message: FALSE

library(dplyr)
library(kableExtra)

# Fit multivariate regression
model <- lm(cbind(Y1, Y2) ~ Z1 + Z2 + Z3 + Z4 + Z5, data = data.df)
s2 <- summary(model)

coef_Y1 <- s2$`Response Y1`$coefficients
coef_Y2 <- s2$`Response Y2`$coefficients

# Convert to data frames (row names as Term column)
coef_Y1_df <- data.frame(Term = rownames(coef_Y1), coef_Y1, row.names = NULL)
coef_Y2_df <- data.frame(Term = rownames(coef_Y2), coef_Y2, row.names = NULL)

coef_Y1_df$Term <- gsub("\\(|\\)", "", coef_Y1_df$Term)
coef_Y2_df$Term <- gsub("\\(|\\)", "", coef_Y2_df$Term)

# Combine side by side including t-values and p-values
coef_combined <- data.frame(
  Term    = coef_Y1_df$Term,
  Y1_Est  = round(coef_Y1_df$Estimate, 2),
  Y1_SE   = round(coef_Y1_df$Std..Error, 2),
  Y1_t    = round(coef_Y1_df$t.value, 3),
  Y1_p    = signif(coef_Y1_df$Pr...t.., 3),
  Y2_Est  = round(coef_Y2_df$Estimate, 2),
  Y2_SE   = round(coef_Y2_df$Std..Error, 2),
  Y2_t    = round(coef_Y2_df$t.value, 3),
  Y2_p    = signif(coef_Y2_df$Pr...t.., 3)
)

# Render table


kable(coef_combined, escape = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE)
```

The output shows the MANOVA tableau for the multivariate model. The results indicate that $Z_1$ and $Z_2$ have significant effects on the dependent variables, suggesting they should be included in the model. The other predictors did not show a significant multivariate effect.

```{r}
#| warning: false
#| message: FALSE
library(car)
Manova(model)
```

We will now construct a 95% prediction ellipse for the new observation in the multivariate model. However, first we calculate the prediction interval for $Y_2$ alone in order to compare the ellipse with the univariate intervals. @tbl-mod2-predint, below, shows this prediction interval.

```{r}
#| echo: false
#| label: tbl-mod2-predint
#| tbl-cap: "95% Prediction Interval for Y2 of new observation"
model.Y2 <- lm(Y2 ~ Z1 + Z2 + Z3 + Z4 + Z5, data = data.df)

prediction.Y2 <- predict(model.Y2, new.df,interval = "prediction", level = 0.95)
kable(prediction.Y1)
```

We now move on to constructing the prediction ellipse for $(Y_1, Y_2)$. The resulting ellipse is illustrated in @fig-pred-ell. We observe that the prediction itself corresponds to the univariate predictions, which is expected since the regression coefficients are the same in the multivariate model.

However, we immediately notice that the extreme points of the ellipse along the $Y_1$ and $Y_2$ axes do not correspond to the boundaries of the respective univariate prediction intervals. The confidence ellipse accounts for the correlation between the outputs, which in this case is very strong ($r = 0.976$). As a result, the ellipse is much more constrained than the rectangular region defined by the univariate intervals. This is clearly visible in the ellipse: regions where one variable is large while the other is small are not included, reflecting the strong positive correlation between $Y_1$ and $Y_2$.
```{r}
#| echo: false
#| fig-cap: "95% prediction ellipse for (Y1, Y2)"
#| label: fig-pred-ell
#| fig-pos: "H"
prediction <- predict(model, new.df) 
confidenceEllipse <- function(mod, newdata, level = 0.95, ggplot = TRUE){
  # labels
  lev_lbl <- paste0(level * 100, "%")
  resps <- colnames(mod$coefficients)
  title <- paste(lev_lbl, "confidence ellipse for", resps[1], "and", resps[2])
  
  # prediction
  p <- predict(mod, newdata)
  
  # center of ellipse
  cent <- c(p[1,1],p[1,2])
  
  # shape of ellipse
  Z <- model.matrix(mod)
  Y <- mod$model[[1]]
  n <- nrow(Y)
  m <- ncol(Y)
  r <- ncol(Z) - 1
  S <- crossprod(resid(mod))/(n-r-1)
  
  # radius of circle generating the ellipse
  # see Johnson and Wichern (2007), p. 399
  tt <- terms(mod)
  Terms <- delete.response(tt)
  mf <- model.frame(Terms, newdata, na.action = na.pass,
                    xlev = mod$xlevels)
  z0 <- model.matrix(Terms, mf, contrasts.arg = mod$contrasts)
  rad <- sqrt((m*(n-r-1)/(n-r-m)) * qf(level,m,n-r-m) *
                z0 %*% solve(t(Z)%*%Z) %*% t(z0))
  
  # generate ellipse using ellipse function in car package
  ell_points <- car::ellipse(center = c(cent), shape = S,
                             radius = c(rad), draw = FALSE)
  
  # ggplot2 plot
  if(ggplot){
    ell_points_df <- as.data.frame(ell_points)
    ggplot2::ggplot(ell_points_df, ggplot2::aes(.data[["x"]], .data[["y"]])) +
      ggplot2::geom_path() +
      ggplot2::geom_point(ggplot2::aes(x = .data[[resps[1]]], 
                                       y = .data[[resps[2]]]),
                          data = data.frame(p)) +
      ggplot2::labs(x = resps[1], y = resps[2],
                    title = title)
  } else {
    # base R plot
    plot(ell_points, type = "l",
         xlab = resps[1], ylab = resps[2],
         main = title)
    points(x = cent[1], y = cent[2])
  }
}
confidenceEllipse(mod = model, newdata = new.df)

```

# Part 2: James–Stein estimator

In this section, we work with the James–Stein (JS) estimator using a dataset of genes. We begin by standardizing the gene expression values.

## 1)

```{r}
#| echo: false
load("Lab3.Rdata")
dimx <- dim(xx) #genes are on columns
m1 <- apply(xx,2,'mean')
xx1 <- scale(xx) 

xx1 <- sweep(xx1, 2, m1, '+') # Add column means back
m2 <- apply(xx1, 2, mean)

xx1 <- xx1 - 10
m3 <- apply(xx1, 2, mean)


```
@fig-hist-m1, below, illustrates the histogram of the average gene expressions before and after standardization. The relative distribution is preserved even after standardizing.

```{r}
#| echo: false
#| fig-cap: "Histogram of average gene expressions before and after standardization"
#| label: fig-hist-m1
#| fig-pos: "H"
# Plot all three histograms in one row
par(mfrow = c(1,3))  # 1 row, 3 columns

hist(m1, main = "Histogram of Column Means", xlab = "Means", col = "lightblue")
hist(m2, main = "After Standardization + Col. Means", xlab = "Means", col = "lightgreen")
hist(m3, main = "After Subtracting 10", xlab = "Means", col = "lightpink")

par(mfrow = c(1,1))  # Reset layout
```

## 2)

We now use the first 5 individuals to estimate the mean vector of the standardized expressions. We compare the MLE against two JS estimators: JS shrinkage toward zero and JS shrinkage toward the grand mean. The formulas for these estimators are defined as follows:

$$
\hat\mu_{JS}=\left(1-\frac{(p-2)\sigma^2}{\|\mathbf Y\|^2}\right)
$$

where

$$
\hat\mu_{JS_{gmean}}=(1-d)\mathbf Y+d\bar{\mathbf Y}
$$

where

$$
d=\frac{p-3}{(p-1)n\sigma^2}
$$

If $\sigma^2$ is unknown, we typically use the MLE estimate of the variance. In this case, we are estimating a mean vector, so the relevant estimator is the variance of the sample mean (MLE). @tbl-sq-error, below, illustrates the squared errors for the three estimators. We observe that both JS estimators perform better than the MLE, with shrinkage toward the grand mean performing the best.

```{r}
#| label: tbl-sq-error
#| echo: false
#| caption: "Squared Errors of the Three Estimators"
n <- dim(xx1)[1]   # number of individuals (rows)
p <- dim(xx1)[2]   # number of genes (columns)
X1  <- xx1[1:5,] 
X2  <- xx1[6:n,]

mle   <- apply(X1,2,'mean') #MLE from first 5 rows
tmean <- apply(X2,2,'mean') #Test mean from remaining rows
JS1 <- (1-(p-2)/sum((mle^2))/5)*mle
d   <- (p-3)/(p-1)/5/var(mle)
JS2 <-(1-d)*mle+d*mean(mle)


SE_MLE  <- sum((tmean - mle)^2)
SE_JS1  <- sum((tmean - JS1)^2)
SE_JS2  <- sum((tmean - JS2)^2)

# Create a table
sq_error_table <- data.frame(
  Estimator = c("MLE", "JS1 (toward 0)", "JS2 (toward mean)"),
  Squared_Error = c(SE_MLE, SE_JS1, SE_JS2)
)

# Print as kable
kable(sq_error_table, caption = "Squared Errors of the Three Estimators") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

@fig-est-scatter, below, plots the estimators against the true mean. The figure is quite crowded, making it difficult to discern all patterns, but the JS estimators appear to perform better in several regions where the MLE performs poorly, such as for the largest means. However, it is not easy to observe any apparent differences between the two JS estimators.

```{r}
#| echo: false
#| fig-cap: "Scatter plot of the three estimators against the mean"
#| label: fig-est-scatter
#| fig-pos: "H"
# Base plot for MLE (black circle outline)
plot(tmean ~ mle, xlab="Estimator", ylab="True Mean", main="Estimator vs True Mean",
     pch=1, col='#000000', cex=1.3, lwd=2)  # larger size and thicker outline
abline(a=0, b=1, lty=2)

# JS1 points (triangle outline, vivid orange)
points(tmean ~ JS1, col='#E69F00', pch=2, cex=1.3, lwd=2)

# JS2 points (square outline, vivid blue)
points(tmean ~ JS2, col='#56B4E9', pch=0, cex=1.3, lwd=2)

# Legend with matching shapes and colors
legend("topleft", legend=c("MLE","JS1 (toward 0)","JS2 (toward mean)"),
       col=c('#000000','#E69F00','#56B4E9'), pch=c(1,2,0), pt.cex=1.3, pt.lwd=2)
```

We now repeat the process using sample sizes of 10, 20, and 40 instead of 5. @tbl-sq-error-diff, below, shows the resulting squared errors. As the sample size increases, the advantage of the JS estimators over the MLE diminishes, and the MSEs become more similar. This is expected, as the shrinkage factor approaches 1 as $n$ increases, causing the JS estimator to converge to the MLE. Furthermore, JS shrinkage is most useful when the MLE has high variance, which occurs for small samples. As the sample size grows and the variance decreases, there is little benefit from introducing bias.
\newpage

```{r}
#| label: tbl-sq-error-diff
#| echo: false
#| message: false
#| warning: false

# Define number of individuals to use for estimating the mean
n_indivs <- c(10, 20, 40)

# Prepare empty list to store squared errors
sq_error_list <- list()

for(n_use in n_indivs){
  
  X1  <- xx1[1:n_use,] 
  X2  <- xx1[(n_use+1):n,]
  
  mle   <- apply(X1,2,'mean')  # MLE from first n_use rows
  tmean <- apply(X2,2,'mean')  # True mean from remaining rows
  
  # JS1 (toward 0)
  JS1 <- (1-(p-2)/sum(mle^2)/n_use)*mle
  
  # JS2 (toward grand mean)
  d   <- (p-3)/(p-1)/n_use/var(mle)
  JS2 <- (1-d)*mle + d*mean(mle)
  
  # Squared errors
  SE_MLE  <- sum((tmean - mle)^2)
  SE_JS1  <- sum((tmean - JS1)^2)
  SE_JS2  <- sum((tmean - JS2)^2)
  
  # Store in a dataframe with n_use
  sq_error_list[[as.character(n_use)]] <- data.frame(
    N_Individuals = n_use,
    Estimator = c("MLE", "JS1 (toward 0)", "JS2 (toward mean)"),
    Squared_Error = c(SE_MLE, SE_JS1, SE_JS2)
  )
}

# Combine all into one table
sq_error_table <- do.call(rbind, sq_error_list)

# Print table with caption inside kable()
kable(sq_error_table, caption = "Squared Errors of the Three Estimators for Different Sample Sizes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

@fig-est-scatter-diff, below, illustrates the estimates for different sample sizes. The same effect is clearly visible in this figure.

```{r}
#| echo: false
#| fig-cap: "Scatter plots of the three estimators against the mean for different sample sizes"
#| label: fig-est-scatter-diff
#| fig-pos: "H"
# Define number of individuals (excluding 5)
n_indivs <- c(10, 20, 40)

# Set up plotting area (2x2 grid; last cell will be empty)
par(mfrow=c(2,2), mar=c(4,4,3,1))

for(n_use in n_indivs){
  
  X1  <- xx1[1:n_use,] 
  X2  <- xx1[(n_use+1):n,]
  
  mle   <- apply(X1,2,'mean')  # MLE from first n_use rows
  tmean <- apply(X2,2,'mean')  # True mean from remaining rows
  
  # JS1 (toward 0)
  JS1 <- (1-(p-2)/sum(mle^2)/n_use)*mle
  
  # JS2 (toward grand mean)
  d   <- (p-3)/(p-1)/n_use/var(mle)
  JS2 <- (1-d)*mle + d*mean(mle)
  
  # Base plot for MLE
  plot(tmean ~ mle, xlab="Estimator", ylab="True Mean", 
       main=paste(n_use, "Individuals"), pch=1, col='#000000', cex=1.2, lwd=2)
  abline(a=0, b=1, lty=2)
  
  # JS1 points (triangle outline, orange)
  points(tmean ~ JS1, col='#E69F00', pch=2, cex=1.2, lwd=2)
  
  # JS2 points (square outline, blue)
  points(tmean ~ JS2, col='#56B4E9', pch=0, cex=1.2, lwd=2)
  
  # Add legend only to first plot in the grid
  if(n_use == 10){
    legend("topleft", legend=c("MLE","JS1 (toward 0)","JS2 (toward mean)"),
           col=c('#000000','#E69F00','#56B4E9'), pch=c(1,2,0), pt.cex=1.2, pt.lwd=2)
  }
}

# Reset plotting parameters to default
par(mfrow=c(1,1))

```

# Part 3: Simulation study: model selection

In this final part, we perform a simulation study on model selection. The model uses a design matrix $Z$ of dimension $1000 \times 950$ with i.i.d. entries $N(0, 1/1000)$. We then generate a response vector according to the linear model

$$
Y=Z\beta+\epsilon,\;\;\epsilon\sim(0,I_{1000})
$$

with $\beta=(3,3,3,3,3,0,\dots,0)^\top$. We then estimate linear regression models of the following model orders:

-   the first 2 variables

-   the first 5 variables

-   the first 10 variables

-   the first 100 variables

-   the first 500 variables

-   all 950 variables

We compare the resulting models with each other, as well as which ones are selected based on AIC and BIC.

We begin by analyzing a single replicate. @tbl-pe, below, plots the conditional expected prediction error (given $\hat\beta$) against estimates of the prediction error using RSS (with known vs. unknown $\sigma^2$), as well as with leave-one-out cross-validation (LOOCV). We observe that all PE estimates would select the true model if used as the sole criterion. The prediction error estimated with known $\sigma^2$ performs the best and behaves relatively well for all $k$. However, in practice, the true $\sigma^2$ is usually unknown. For the other two methods, the situation is slightly more complicated: for smaller values of $k$, the LOOCV estimator often performs better, although the difference is small. For larger values of $k$, performance deteriorates, particularly for $k = 950$.

```{r}
#| echo: false
#| tbl-cap: "Comparison of True PE and Estimated Prediction Errors"
#| label: tbl-pe
set.seed(2025)
rm(list = ls())

library(knitr)

# -----------------------------
# Parameters
# -----------------------------
n <- 1000
p <- 950
rep1 <- 100
k_values <- c(2,5,10,100,500,950)
nk <- length(k_values)

# -----------------------------
# Single replicate setup
# -----------------------------
# Preallocate
PE <- rep(0, nk)
PRED1 <- rep(0, nk)
PRED2 <- rep(0, nk)
PRED3 <- rep(0, nk)

# True beta and mean
beta <- rep(0,p)
beta[1:5] <- 3
Z <- matrix(rnorm(n*p,0,1/sqrt(n)), nrow=n)
mu <- Z %*% beta

# Single replicate
epsilon <- rnorm(n)
Y <- mu + epsilon

# Loop over models
i <- 1
for (k in k_values) {
  Zn <- Z[,1:k]
  M1 <- solve(t(Zn) %*% Zn) %*% t(Zn)
  M2 <- Zn %*% M1
  betahat <- M1 %*% Y
  muhat <- Zn %*% betahat
  
  # PE
  PE[i] <- sum((muhat - mu)^2) + n
  
  # RSS and predictions
  RSS <- sum((Y - muhat)^2)
  PRED1[i] <- RSS + 2*k
  PRED2[i] <- RSS*(1 + 2*k/(n-k))
  
  # LOOCV PE
  res1 <- (Y - muhat)/(1 - diag(M2))
  PRED3[i] <- sum(res1^2)
  
  i <- i + 1
}

# -----------------------------
# AIC and BIC for single replicate
# -----------------------------

epsilon_single <- rnorm(n)
Y_single <- mu + epsilon_single

AIC_vals <- numeric(nk)
BIC_vals <- numeric(nk)

i <- 1
for (k in k_values) {
  Zn <- Z[,1:k]
  M1 <- solve(t(Zn) %*% Zn) %*% t(Zn)
  betahat <- M1 %*% Y_single
  muhat <- Zn %*% betahat
  
  RSS <- sum((Y_single - muhat)^2)
  
  AIC_vals[i] <- n*log(RSS/n) + 2*k
  BIC_vals[i] <- n*log(RSS/n) + k*log(n)
  i <- i + 1
}

# Differences between estimated PE and true PE
Diff_PRED1 <- PRED1 - PE
Diff_PRED2 <- PRED2 - PE
Diff_PRED3 <- PRED3 - PE

# Identify the selected model for each criterion
k_min_PRED1 <- k_values[which.min(PRED1)]
k_min_PRED2 <- k_values[which.min(PRED2)]
k_min_PRED3 <- k_values[which.min(PRED3)]
k_min_PE    <- k_values[which.min(PE)]



# Tabulate all results
PE_comparison <- data.frame(
  k = k_values,
  True_PE = PE,
  PRED1 = PRED1,
  PRED2 = PRED2,
  PRED3 = PRED3,
  Diff_PRED1 = Diff_PRED1,
  Diff_PRED2 = Diff_PRED2,
  Diff_PRED3 = Diff_PRED3
)

library(knitr)
kable(PE_comparison, digits=2)


```

@tbl-aic-bic, below, illustrates the AIC and BIC for each model. We see that the BIC criterion results in the correct model order being chosen, while AIC selects the model with $k=950$. The reason for this is that RSS will never increase when adding more predictors to the model. When many predictors are added, RSS becomes very small, and the log-likelihood term in AIC dominates the penalty term. Because AIC has a relatively weak penalty (2k), it favors the largest model even though many predictors are unnecessary. BIC, on the other hand, uses a stronger penalty that grows with $\log(n)$, which prevents it from overfitting and correctly identifies the true model order.

```{r}
#| echo: false
#| tbl-cap: "AIC and BIC values for each model"
#| label: tbl-aic-bic
AIC_choice <- k_values[which.min(AIC_vals)]
BIC_choice <- k_values[which.min(BIC_vals)]

AIC_BIC_table <- data.frame(
  k = k_values,
  AIC = AIC_vals,
  BIC = BIC_vals
)

kable(AIC_BIC_table, digits = 2)

```

```{r}
#| echo: false

# -----------------------------
# Multi-replicate setup
# -----------------------------
# Preallocate matrices
PE_mat <- matrix(0, nrow=rep1, ncol=nk)
PRED1_mat <- matrix(0, nrow=rep1, ncol=nk)
PRED2_mat <- matrix(0, nrow=rep1, ncol=nk)
PRED3_mat <- matrix(0, nrow=rep1, ncol=nk)

# Precompute inverses and hat diagonals
XtX_inv_list <- vector("list", nk)
H_diag_list <- vector("list", nk)

for (i in seq_along(k_values)) {
  k <- k_values[i]
  Zn <- Z[, 1:k]
  
  XtX_inv <- solve(crossprod(Zn))
  XtX_inv_list[[i]] <- XtX_inv
  H_diag_list[[i]] <- rowSums((Zn %*% XtX_inv) * Zn)
}

# Matrices to store AIC/BIC counts
AIC_choice_counts <- integer(nk)
BIC_choice_counts <- integer(nk)

# Main loop over repetitions
for (j in 1:rep1) {
  epsilon <- rnorm(n)
  Y <- mu + epsilon
  
  AIC_vals <- numeric(nk)
  BIC_vals <- numeric(nk)
  
  for (i in seq_along(k_values)) {
    k <- k_values[i]
    Zn <- Z[,1:k]
    XtX_inv <- XtX_inv_list[[i]]
    h <- H_diag_list[[i]]
    
    # Least squares
    betahat <- XtX_inv %*% crossprod(Zn, Y)
    muhat <- Zn %*% betahat
    
    # Conditional PE
    PE_mat[j,i] <- sum((muhat - mu)^2) + n
    
    # RSS and predictions
    RSS <- sum((Y - muhat)^2)
    PRED1_mat[j,i] <- RSS + 2*k
    PRED2_mat[j,i] <- RSS*(1 + 2*k/(n-k))
    
    # LOOCV PE
    res1 <- (Y - muhat)/(1 - h)
    PRED3_mat[j,i] <- sum(res1^2)
    
    # AIC/BIC for this replicate
    AIC_vals[i] <- n*log(RSS/n) + 2*k
    BIC_vals[i] <- n*log(RSS/n) + k*log(n)
  }
  
  # Count selected models
  AIC_choice_counts[which.min(AIC_vals)] <- AIC_choice_counts[which.min(AIC_vals)] + 1
  BIC_choice_counts[which.min(BIC_vals)] <- BIC_choice_counts[which.min(BIC_vals)] + 1
}


```

Having examined a single replicate, we then performed 100 simulations of the output vectors to study how the prediction errors and model selection criteria behave across different model orders. We first consider the differences between the estimated and true prediction errors, illustrated with boxplots in @fig-pe-diff, below. The first two estimators appear approximately unbiased for all model orders. In the case of the first estimator, which uses the true $\sigma^2$, the variance remains roughly constant across model orders, as expected since the only source of randomness is the residual. For the estimator using the estimated $\sigma^2$, the bias still appears negligible, but the variance increases substantially for higher model orders. Finally, LOOCV clearly overestimates the expected prediction error for the higher model orders and exhibits considerably more variability.

```{r}
#| echo: false
#| fig-cap: "Boxplots of cPE minus PE for three prediction-error estimators across model orders k."
#| label: fig-pe-diff
#| fig-pos: "H"
par(mfrow = c(1,3), las = 2)  # las=2 => vertical axis labels

boxplot(PRED1_mat - PE_mat,
        main = "PRED1 − PE",
        xlab = "Model order k",
        ylab = "cPE − PE",
        names = k_values)

boxplot(PRED2_mat - PE_mat,
        main = "PRED2 − PE",
        xlab = "Model order k",
        ylab = "cPE − PE",
        names = k_values)

boxplot(PRED3_mat - PE_mat,
        main = "PRED3 − PE",
        xlab = "Model order k",
        ylab = "cPE − PE",
        names = k_values)

par(mfrow = c(1,1), las = 0)
```

However, since we may wish to use the estimated prediction error to select the model order, it is also of interest to examine how much it varies across different simulations. Because the prediction error is consistently much larger for higher model orders, it is most informative to focus on the distribution for the lower model orders. Boxplots for this are illustrated in @fig-pe, below. In the case of the true conditional error, there is some overlap between the distributions for model orders 5 and 10, while the expected error for model order 2 is larger, which is natural because we are not using all the available information. The expected error for $k=10$ can still be small as long as the additional coefficients beyond $k=5$ are estimated reasonably close to 0. For the estimated prediction errors, there is considerable overlap across all model orders, with $k=5$ and $k=10$ being particularly close.

```{r}
#| echo: false
#| fig-cap: "Boxplots of true and estimated prediction errors for different model orders k; PRED1 uses true sigma2, PRED2 uses estimated sigma2, and PRED3 is LOOCV."
#| label: fig-pe
# Set up 2x2 plotting area
par(mfrow = c(2,2), mar = c(4,4,3,1))  # bottom, left, top, right margins

# Define k values for each subset
k_PE    <- k_values[1:3]
k_PRED1 <- k_values[1:4]
k_PRED2 <- k_values[1:4]
k_PRED3 <- k_values[1:4]

# Plot PE
boxplot(PE_mat[,1:3], names = k_PE, xlab = "Model order k", ylab = "Prediction Error", 
        main = "PE")

# Plot PRED1
boxplot(PRED1_mat[,1:4], names = k_PRED1, xlab = "Model order k", ylab = "Estimated PE", 
        main = "PRED1")

# Plot PRED2
boxplot(PRED2_mat[,1:4], names = k_PRED2, xlab = "Model order k", ylab = "Estimated PE", 
        main = "PRED2")

# Plot PRED3
boxplot(PRED3_mat[,1:4], names = k_PRED3, xlab = "Model order k", ylab = "Estimated PE", 
        main = "PRED3")

# Reset plotting parameters to default
par(mfrow = c(1,1))

```

However, although there is considerable overlap between the prediction error estimates across different model orders, this does not necessarily imply a high risk of selecting the wrong model in most simulations. The reason is that the error estimates for different model orders are not independent: if one model has a particularly high or low estimate in a replicate, the other estimates are affected in a correlated way. As a result, the relative ordering of the estimates often still favors the correct model, even when the absolute values appear overlapping. We can define an incorrect decision either as choosing the model with the wrong model order or as not choosing the model with the lowest expected prediction error (which will typically be the true model, although in rare cases a different model may perform slightly better due to random fluctuations in the data). @tbl-wrong-decisions, below, illustrates that we would indeed select the wrong model far less often than the overlapping distributions suggest, but still often enough to show that using the prediction error estimate alone is not a guarantee for choosing the correct model, even in a very simple setup like ours.

```{r}
#| echo: false
#| tbl-cap: "Number of incorrect model selections over 100 simulations"
#| label: tbl-wrong-decisions

library(knitr)

# Assume true model is k = 5
true_index <- which(k_values == 5)

# Wrong model order for each estimator
wrong_model_PRED1 <- sum(apply(PRED1_mat, 1, which.min) != true_index)
wrong_model_PRED2 <- sum(apply(PRED2_mat, 1, which.min) != true_index)
wrong_model_PRED3 <- sum(apply(PRED3_mat, 1, which.min) != true_index)

# Wrong minimum PE choice
min_true_PE_index <- apply(PE_mat, 1, which.min)
wrong_PEchoice_PRED1 <- sum(apply(PRED1_mat, 1, which.min) != min_true_PE_index)
wrong_PEchoice_PRED2 <- sum(apply(PRED2_mat, 1, which.min) != min_true_PE_index)
wrong_PEchoice_PRED3 <- sum(apply(PRED3_mat, 1, which.min) != min_true_PE_index)

# Create a summary table
error_table <- data.frame(
  Estimator = c("PRED1", "PRED2", "PRED3"),
  Wrong_Model_Order = c(wrong_model_PRED1, wrong_model_PRED2, wrong_model_PRED3),
  Wrong_Min_PE = c(wrong_PEchoice_PRED1, wrong_PEchoice_PRED2, wrong_PEchoice_PRED3)
)

# Display the table
kable(error_table, digits = 0)
```

Finally, it is also of interest to see how many times each model order was selected using AIC or BIC. This is illustrated in @tbl-aic-bic-counts, below. We see that BIC picks the correct model order in the majority of cases. However, when it does not select the correct model, it tends to choose the more conservative $k=2$, indicating a tendency to underfit, which should be kept in mind. On the other hand, AIC consistently selects the highest possible model order, making it clear that using AIC as a selection criterion is not advisable.
```{r}
#| tbl-cap: "Number of times each model was selected by AIC and BIC over 100 replicates"
#| label: tbl-aic-bic-counts
#| echo: false
# -----------------------------
# Table for AIC/BIC model selections over 100 replicates using kable
# -----------------------------
selection_table <- data.frame(
  Model = k_values,
  AIC_Selected = AIC_choice_counts,
  BIC_Selected = BIC_choice_counts
)

kable(selection_table)

```

# Use of AI

AI (ChatGPT) has been utilized in this report primarily to assist with coding, including constructing and formatting plots, generating LaTeX code, and suggesting functions. It was also used for light spelling and grammar checks, and for suggestions on how to more clearly convey results.
