---
title: "Computer Lab 2: High Dimensional Data Analysis"
author: "Sara Kopelman"
date: today
date-format: long
format:
  pdf:
    toc: false
    number-sections: false
    booktabs: true
header-includes:
  - \usepackage{float}  
---

# Printing Banknotes

## 1-2)

We will begin by plotting the data. @tbl-colmeans, [-@tbl-var] and [-@tbl-cor] describes the mean, variance and correlations in the dataset. We note that the average left height is slightly larger than the right, and that the average distance of the inner frame to the bottom is slightly smaller than that to the top, although we have not performed any tests to determine if these differences are statistically significant. We also note that the distance to the bottom and the diagonal has the largest variance. We also see that there are several moderately large correlations present in the dataset, which is natural given the nature of the variables.

```{r}
#| echo: false
#| results: 'hide'
#| warning: false

df <- read.csv("banknotes.csv")
BG <- df[,1:6]

library(knitr)
library(kableExtra)

head(BG)




```

```{r}
#| echo: false
#| label: tbl-colmeans
# 2. Column means
kable(t(colMeans(BG)),
      caption = "Column Means of BG",
      digits = 4) %>%
  kable_styling(full_width = FALSE)


```

\newpage

```{r}
#| echo: false
#| label: tbl-var
# 3. Variance matrix
kable(var(BG), 
      caption = "Variance-Covariance Matrix of BG",
      digits = 4) %>%
  kable_styling(full_width = FALSE)


```

```{r}
#| echo: false
#| label: tbl-cor
# 4. Correlation matrix
kable(cor(BG),
      caption = "Correlation Matrix of BG",
      digits = 4) %>%
  kable_styling(full_width = FALSE)
```

@fig-banknotes_pairs, on the following page, plots the variables in the dataset against each other. We once again note that there are correlations visible between several variable pairs (such as for example *right* and *left*), and we also see the presence of some partially separated groups in the data.

```{r}
#| echo: false
#| fig-cap: "Scatterplot matrix of the variables in the banknotes dataset"
#| label: fig-banknotes_pairs
#| fig-pos: "H"
plot(BG)
```

Furthermore, @fig-banknotes_qq, on the following page, illustrates Normal QQ-plots for the variables in the dataset. We observe that *length, right, top* and *left* adhere quite well to the normal distribution. However, *bottom* looks quite light-tailed while *diagonal* looks possible bi-modal.

```{r}
#| echo: false
#| fig-cap: "QQ-plots of the variables in the banknotes dataset"
#| label: fig-banknotes_qq
#| fig-pos: "H"
par(mfcol = c(2, 3))

for (i in 1:6) {
  qqnorm(BG[, i],
         main = paste("QQ Plot of", colnames(BG)[i]),
         pch = 19, cex = 0.6)
  qqline(BG[, i], col = "red", lwd = 2)
}

```

## 3)

The below output shows how to compute the Hotellings $T^2$ statistic in two difference ways. For a sample of size $n$ with sample mean-vector $\bar X$ (with dimension $p\times1$) and sample covariance matrix $S$ the statistic is defined as follows:

$$
T^2=n(\bar X-\mu_0)^\top S^{-1}(\bar X-\mu_0)
$$

where $\mu_0$ is the mean vector under the null-hypothesis.

```{r}
n <- dim ( BG ) [1]
p <- dim ( BG ) [2]
S <- cov ( BG )
IS <- solve ( S )
barX <- colMeans ( BG )

a <- (n -1) * p / (n - p ) * qf (0.95,p,n-p)
q <- 3
mu0 <- c(210 ,130 ,130 ,10 ,10 ,140)
t2 <- n * ( barX - mu0 ) %*% IS %*% ( barX - mu0 )
cx <- as.matrix ( barX - mu0 )
T2 <- n * t ( cx ) %*% IS %*% cx
print(t2)
print(T2)
print(a)
```

Hotelling's $T^2$ test is used to determine whether the mean vector of a multivariate dataset differs significantly from a hypothesized mean vector. In this analysis, the data matrix BG has `n` rows (observations) and `p` columns (variables). We first compute the sample covariance matrix `S` of the data and its inverse, `IS = solve(S)`.

The sample mean vector `barX` is calculated as the column-wise mean of BG. The hypothesized mean vector `mu0` represents the null hypothesis. The Hotelling's $T^2$ statistic is then computed as according to the formula above in two different ways, either by directly multiplying the difference-vector with the inverse, or by converting it into a matrix and transposing when needed to for the dimensions to align.

As can be seen both ways of calculation results in the same value for $T^2$, 41750.76. One may, however, observe that only the second method transposes the difference-vector, yet the code still runs correctly for both. The reason for this stems from the difference in R in how vectors are handled vs matrices. The `%*%` operator automatically treats a vector as a column or row vector depending on the context, opting for the configuration that would allow for the operation to be performed. *Matrices*, however need to have compatible dimensions for operations to run. As the `cx` object is a matrix it needs to be transposed for the operation to be performed, which is done when calculating `T2` . In summary, vectors and one-dimensional matrices are different types of objects in R.

$T^2$ can be converted to an F-statistic using `F = (n - p) / (p * (n - 1)) * T2`. The critical value `a = (n - 1) * p / (n - p) * qf(0.95, p, n - p)` corresponds to the 95% confidence level. If $T^2$ exceeds this the sample mean differs significantly from the hypothesized mean, accounting for the correlation structure of the variables. The observed statistic does exceed the critical value and we can thus reject the null.

## 4-5)

The following output provides a function to calculate the that verifies if a point lies inside the six-dimensional ellipsoid serving as the 95% confidence region for the mean vector of bank notes based on Hotelling’s $T^2$ statistic.

```{r}
InHotCR=function(mu0,barX,IS,n){ 
  mu0=as.vector(mu0)   #mu0 in the code below
  barX=as.vector(barX) # must be a vector not a matrix!
  p=dim(IS)[1]
  a=(n-1)*p/(n-p)*qf(0.95,p,n-p)
  cx=as.matrix(barX-mu0)
  T2=n*t(cx)%*%IS%*%cx
  InHotCR=T2<a
  InHotCR}
```

Using this function we test new production line with a new mean vector. The test statistic however shows that the new dimensions are not comparable to the old ones.

```{r}
m0=c(214.97,130,129.67,8.3,10.16,141.52)
InHotCR(m0,barX,IS,n)
```

## 6)

The following function a new mean vector falls within Bonferroni’s confidence rectangular region for the mean value of the old bank note dimensions. As we see from the result the new vector does not fall within this region either.

```{r}
InBonCR=function(mu0,barX,S,n){ 
  mu0=as.vector(mu0)  #mu0 must be a vector not a matrix!
  barX=as.vector(barX)
  p=dim(S)[1]
  alpha=0.05/p
  a=qf(1-alpha,1,n-1)*diag(S)/n
  cx=as.matrix(barX-mu0)
  T2=cx^2
  T2<a
  InBonCR=T2<a
  InBonCR
}

prod(InBonCR(m0,barX,S,n))==1
```

## 7-8)

Figure @fig-banknotes_1dregion, below, shows the Bonferroni and Hotelling confidence intervals for each of the six variables, centered at the mean of the old production line. The red points indicate the endpoints of the Bonferroni intervals, while the blue vertical ticks show the Hotelling marginal intervals. The black stars represent the mean vector of the new production line.

We observe that the new production line’s mean lies outside the Bonferroni intervals for all variables, indicating that the new line may differ significantly from the old line in those aspects. The variable length, however, falls within the Hotelling marginal interval and is right at the edge of the Bonferroni interval, suggesting that its difference is less pronounced.

```{r}
#| echo: false
#| fig-cap: "Bonferroni and Hotelling confidence intervals for each variable with null mean highlighted"
#| label: fig-banknotes_1dregion
#| fig-pos: "H"
par(mfcol=c(2,3))

alpha=0.05/p
a=sqrt(qf(1-alpha,1,n-1)*diag(S)/n)
aa=sqrt((n-1)*p/(n-p)*qf(0.95,p,n-p)*diag(S)/n)
LP=barX-a
RP=barX+a
LPP=barX-aa
RPP=barX+aa
varnames <- colnames(BG)

for(i in 1:6){
  mn = min(LP[i], LPP[i])
  mx = max(RP[i], RPP[i])

  plot(c(LP[i], RP[i]), c(0,0),
       col='red', yaxt='n',
       xlim=c(mn, mx),
       main = varnames[i])   # now it appears

  abline(0,0)
  points(m0[i], 0, pch="*", cex=2, col="black")
  points(c(barX[i]-aa[i], barX[i]+aa[i]), c(0,0),
         col='blue', pch="|")
}


```

@fig-banknotes_2dregion, below, plots the pairwise 2D Hotelling confidence region for the first variable pair (plots for the other variable pairs are available in the .qmd document but were omitted from the report for clarity). We can now see why the mean values of the banknote dimensions from the new production line fail to lie within Hotelling’s confidence region, even though the mean *length* by itself falls within its 1D interval. Specifically, while the mean *length* lies within the Hotelling interval, the mean of the *left* dimension falls outside its axis, so the observed *length–left* pair ultimately falls outside the 2D region. The same reasoning applies to the full 6-dimensional mean vector, which would need to lie inside the boundaries of the 6-dimensional ellipsoid to be fully consistent with the old line.

```{r}
#| label: fig-banknotes_2dregion
#| echo: false
#| fig-cap: "Pairwise 2D Hotelling confidence region for the first variable pair."
#| fig-pos: "H"
#| warning: false

library(ellipse)

varnames <- colnames(BG)

i <- 1
j <- 2
SS <- S[c(i,j), c(i,j)]
mm <- barX[c(i,j)]

ell <- ellipse(SS/n, centre = mm,
               t = sqrt((n-1)*p/(n-p)*qf(0.95,p,n-p)),
               npoints = 1000)

xlim <- range(c(ell[,1], m0[i]))
ylim <- range(c(ell[,2], m0[j]))

# Base R plot directly
plot(ell, type='l', col = "blue",
     xlim = xlim, ylim = ylim,
     main = paste(varnames[i], "vs", varnames[j]),
     asp=1)
points(m0[i], m0[j], pch="*", col="black", cex=1.5)
rect(LP[i], LP[j], RP[i], RP[j], border="red")
```

```{r}
#| echo: false
#| fig-show: 'asis' 
#| warning: false
varnames <- colnames(BG)

in_report <- knitr::is_latex_output() || knitr::is_html_output()

for(i in 1:5){
  for(j in (i+1):6){
    SS <- S[c(i,j), c(i,j)]
    mm <- barX[c(i,j)]
    
    ell <- ellipse(SS/n, centre = mm,
                   t = sqrt((n-1)*p/(n-p)*qf(0.95,p,n-p)),
                   npoints = 1000)
    
    xlim <- range(c(ell[,1], m0[i]))
    ylim <- range(c(ell[,2], m0[j]))
    
    # Skip the first pair
    if(!(i==1 && j==2)){
      # Only plot if not in report
      if(!in_report){
        plot(ell, type='l', col = "blue",
             xlim = xlim, ylim = ylim,
             main = paste(varnames[i], "vs", varnames[j]),
             asp=1)
        
        points(m0[i], m0[j], pch="*", col="black", cex=1.5)
        rect(LP[i], LP[j], RP[i], RP[j], border="red")
      }
    }
  }
}


```

## 9)

We evaluate a new production line and tests its mean against the old line. The results are described in the below output. We see that once again the new production line is outside the mean vector. Although individual variables (such as once again *length* for Hotellinger) may fall within their marginal intervals, small but consistent differences in several other dimensions are enough for the overall mean vector to fall outside the 6-dimensional confidence region. This sensitivity reflects the multivariate nature of the test.

```{r}
m1 = c(214.99, 129.95, 129.73, 8.51, 9.96, 141.55)
InHotCR(m1,barX,IS,n)

```

```{r}
InBonCR(m1,barX,S,n)
```

## 10)

After yet another tuning, the mean vector of the new production line shows that *length* now lies within its Bonferroni confidence interval, indicating that this dimension may be adequately aligned with the old line. However, the other dimensions remain outside their respective intervals, and the overall vector is still outside the 6-dimensional Hotelling region. This indicates that while the tuning has improved the alignment of *length*, the production line still requires further adjustments in the remaining dimensions to align with the original banknotes.

```{r}
m2 = c(214.9473, 129.9243, 129.6709, 8.3254, 10.0389, 141.4954)
InHotCR(m2,barX,IS,n)

```

```{r}
InBonCR(m2,barX,S,n)
```

# Simulation

## Simulation 1: Global Null Testing

In this section we will consider the sequence of independent random variables $X_1,\dots,X_p$ such that $X_i\sim N(\mu_i,1)$, under the three following scenarios

-   $\mu_1=1.2\sqrt{2\log p},\mu_2=\cdots=\mu_p=0$

-   $\mu_1=\cdots=\mu_{100}=0.5\sqrt{2\log p},\mu_{101}=\cdots=\mu_p=0$

-   $\mu_1=\cdots=\mu_{1000}=0.2\sqrt{2\log p},\mu_{1001}=\cdots=\mu_p=0$

with $p=5000$. Under these scenarios we will test the global null hypothesis

$$
H_0:\mu_1=\cdots=\mu_p=0
$$

with significance level $\alpha=0.05$, using Bonferroni and chi-square tests by simulating 1000 replicates in order to estimates their power. @fig-multitest-power, below illustrates the estimated power of both tests under each scenario.

```{r}
#| label: fig-multitest-power
#| echo: false
#| fig-cap: "Estimated power of the Bonferroni max-test and the chi-square global test under three signal configurations with p = 5000"
#| fig-pos: "H"
#| warning: false
#| message: false
n <- 5000
rep1 <- 1000
alpha <- 0.05

# Store results
scenarios <- list(
  mu1 = {mu <- rep(0, n); mu[1] <- 1.2*sqrt(2*log(n)); mu},
  mu100 = {mu <- rep(0, n); mu[1:100] <- 0.5*sqrt(2*log(n)); mu},
  mu1000 = {mu <- rep(0, n); mu[1:1000] <- 0.2*sqrt(2*log(n)); mu}
)

powers <- data.frame(Test = character(), Scenario = character(), Power = numeric())

for (name in names(scenarios)) {
  mu <- scenarios[[name]]
  Bon <- rep(0, rep1)
  Chi <- rep(0, rep1)
  
  for (j in 1:rep1) {
    X <- rnorm(n) + mu
    Bon[j] <- max(abs(X)) > qnorm(1 - alpha/(2*n))
    Chi[j] <- sum(X^2) > qchisq(1 - alpha, n)
  }
  
  powers <- rbind(
    powers,
    data.frame(Test = "Bonferroni", Scenario = name, Power = mean(Bon)),
    data.frame(Test = "Chi-squared", Scenario = name, Power = mean(Chi))
  )
}

library(ggplot2)

# Bar plot
ggplot(powers, aes(x = Scenario, y = Power, fill = Test)) +
  geom_bar(stat = "identity", position = "dodge") +
  ylim(0,1) +
  labs(title = "Power comparison: Bonferroni vs Chi-squared",
       y = "Estimated Power") +
  theme_minimal()


```

We observe that while the Bonferroni test is much more powerful than the chi-square test under the first scenario, it is less powerful under the second, and the relationship is completely reversed in the third. To understand why this is, we need to consider how the Bonferroni test works.

The Bonferroni method controls the family-wise error rate by testing each individual hypothesis at level $\alpha/p$. Equivalently, we reject the global null hypothesis $H_0$ if any individual $p$–value satisfies

$$p_j < \frac{\alpha}{p}.$$

Bonferroni thus reacts to the most significant coordinate. In the first scenario, the signal is sparse but strong, meaning that the single non-zero mean produces a p-value above the threshold, resulting in a fairly powerful test. In the later scenarios, however, the signal is less sparse but weaker: although many coordinates deviate from zero, the individual p-values are not extreme enough to pass the Bonferroni threshold, resulting in lower power.

The chi-square test, on the other hand, aggregates evidence across all coordinates using the statistic

$$
T = \sum_{j=1}^p X_j^2,
$$

which is compared to the $\chi^2_p$ distribution under $H_0$. This makes the chi-square test weak in sparse scenarios, even if the signal is strong. However, in the denser scenarios, it outperforms the Bonferroni test because each non-zero signal, although small, contributes to the overall test statistic, accumulating evidence against the null hypothesis.

In summary, the Bonferroni test is most powerful when the signal is concentrated in a few strong coordinates, because it only requires one extremely significant p-value to reject the null. In contrast, the chi-square test performs better when the signal is spread across many coordinates, as it accumulates moderate effects to detect deviations from the null.

## Simulation 2: Multiple Testing

In this section we will explore multiple testing, considering the sequence of independent random variables $X_1,\dots,X_p$ such that $X_i\sim N(\mu_i,1)$, under the following two scenarios

-   $\mu_1=\cdots=\mu_{10}=\sqrt{2\log p},\mu_{11}=\cdots=\mu_p=0$

-   $\mu_1=\cdots=\mu_{500}=\sqrt{2\log p},\mu_{501}=\cdots=\mu_p=0$

with $p=5000$. Under these scenarios we will test the hypotheses

$$
H_{0i}:\mu_i=0
$$

for $i\in\{1,\dots,p\}$ with significance level $\alpha=0.05$, using the Bonferroni and Benjamini-Hochberg multiple testing procedure,by simulating 1000 replicates in order to estimates their family wise error rate (FWER), false discovery rate (FDR), and power. @tbl-multitest, below, describes the estimation for these metric for both scenarios.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: 'asis'
n <- 5000
rep1 <- 1000
alpha <- 0.05

run_simulation <- function(k, n, rep1, alpha) {
  mu <- rep(0, n)
  mu[1:k] <- sqrt(2*log(n))
  
  # Initialize vectors
  T1Bon <- FDPBon <- TPPBon <- numeric(rep1)
  T1BH <- FDPBH <- TPPBH <- numeric(rep1)
  
  qBH <- seq(1, n) * alpha / (2 * n)
  
  for (i in 1:rep1) {
    X <- rnorm(n) + mu
    
    # Bonferroni
    Bon <- abs(X) > qnorm(1 - alpha/(2*n))
    RBon <- sum(Bon)
    VBon <- sum(Bon[(k+1):n])
    T1Bon[i] <- VBon > 0
    FDPBon[i] <- VBon / max(RBon, 1)
    TPPBon[i] <- (RBon - VBon) / k
    
    # BH
    U <- sort(abs(X), decreasing = TRUE, index.return = TRUE)
    kBHs <- 0
    decBH <- which(U$x > qnorm(1 - qBH))
    if (length(decBH) > 0) kBHs <- max(decBH)
    BHs <- rep(0, n)
    BHs[1:kBHs] <- 1
    BH <- rep(0, n)
    BH[U$ix] <- BHs
    
    RBH <- sum(BH)
    VBH <- sum(BH[(k+1):n])
    T1BH[i] <- VBH > 0
    FDPBH[i] <- VBH / max(RBH, 1)
    TPPBH[i] <- (RBH - VBH) / k
  }
  
  # Return summarized results
  list(
    FWERBon = mean(T1Bon),
    FWERBH = mean(T1BH),
    FDRBon = mean(FDPBon),
    FDRBH = mean(FDPBH),
    PowerBon = mean(TPPBon),
    PowerBH = mean(TPPBH)
  )
}

# Run both scenarios
results_k10  <- run_simulation(k = 10, n, rep1, alpha)
results_k500 <- run_simulation(k = 500, n, rep1, alpha)

```

```{r}
#| echo: false
#| label: tbl-multitest
#| warning: false
#| message: false
#| results: 'asis'
library(knitr)
library(dplyr)

# Combine results into a long table
results_table <- data.frame(
  Scenario = rep(c("k = 10", "k = 500"), each = 2),
  Test = rep(c("Bonferroni", "BH"), times = 2),
  FWER = c(results_k10$FWERBon, results_k10$FWERBH,
           results_k500$FWERBon, results_k500$FWERBH),
  FDR   = c(results_k10$FDRBon, results_k10$FDRBH,
            results_k500$FDRBon, results_k500$FDRBH),
  Power = c(results_k10$PowerBon, results_k10$PowerBH,
            results_k500$PowerBon, results_k500$PowerBH)
)

# Round for display
results_table[, 3:5] <- round(results_table[, 3:5], 3)

# Print table
kable(
  results_table,
  caption = "FWER, FDR, and Power for Bonferroni and BH tests across two scenarios",
  align = c("l", "l", "c", "c", "c")
)


```

We observe that while the FWER stays around $\alpha = 0.05$ for Bonferroni in both scenarios, it increases from $0.288$ to $1$ for the Benjamini–Hochberg (BH) procedure.

The BH procedure sorts the $p$-values $p_{(1)}\le\cdots\le p_{(m)}$ and finds the largest index $k$ with

$$
p_{(k)} \le \frac{k}{m}\alpha,
$$

then rejects $H_{(1)},\dots,H_{(k)}$.

In the first scenario, the signal is sparse, so the true nulls—which generally have larger $p$-values—are compared against moderately small thresholds. For example, assuming the 10 non-nulls have the smallest $p$-values and are correctly rejected, the first null is compared to a threshold of $\frac{11}{p}\alpha$, meaning that most nulls are unlikely to be rejected, and the FWER remains low.

In the second scenario, the signal is dense. Here, many true alternatives likely have small $p$-values, which pushes the BH rejection threshold to larger values. Consequently, the nulls are more likely to be compared against these higher thresholds, and the probability of at least one false rejection in each replicate grows toward 1.

On the other hand, the false discovery rate starts at 0.011 for Bonferroni in the first scenario and decreases to 0 in the second, a result of Bonferroni’s conservativeness. Because the Bonferroni threshold is very strict, it produces very few rejections overall, and when it does reject, it almost always corresponds to a true signal. This also means that the power remains roughly the same in both scenarios: although there are more signals to discover in the second scenario, the fraction detected remains low.

In contrast, BH is designed to control the FDR rather than the FWER. This allows it to reject more hypotheses, accepting some false discoveries as long as their proportion is controlled. As a result, in the sparse scenario, power is relatively low, but in the dense scenario, power increases. By allowing a controlled rate of false discoveries, BH can detect a much larger number of true signals overall, which is especially beneficial in dense settings.

# Use of AI

AI (ChatGPT) has been utilized in this report primarily to assist with coding, including constructing and formatting plots, generating LaTeX code, and suggesting functions. It was also used for light spelling and grammar checks, and for suggestions on how to more clearly convey results.
