---
title: "Computer Lab 3: High Dimensional Data Analysis"
author: "Sara Kopelman"
date: today
date-format: long
format:
  pdf:
    toc: false
    number-sections: false
    booktabs: true
header-includes:
  - \usepackage{float}  
---

# Part 1: Monte Carlo simulation

In this section, we simulate data from the linear model

$$
Y = X\beta + \epsilon
$$

where $X$ is an $n \times p$ design matrix drawn from a multivariate normal distribution with independent rows

$$
X_i \sim N_p(0, \Sigma)
$$

and covariance matrix

$$
\Sigma = \rho J_p + (1 - \rho) I_p,
$$

where $J_p$ is the $p \times p$ matrix of ones, $I_p$ is the $p \times p$ identity matrix, and $\rho$ denotes the pairwise correlation among covariates.

The true coefficient vector is given by

$$
\beta_1 = \beta_2 = \cdots = \beta_{10} = 3,
\qquad
\beta_{11} = \cdots = \beta_p = 0,
$$

and the noise term satisfies

$$
\epsilon \sim N_n(0, I_n).
$$

We consider all combinations of

$$
\rho \in \{0.25, 0.50, 0.95\},
\qquad
p \in \{10, 100, 200\},
$$

with a fixed sample size $n = 50$. For each configuration, 200 independent datasets are generated.

For each pair $(\rho, p)$, the simulation steps are:

1.  Construct the covariance matrix\
    $$
    \Sigma = \rho J_p + (1 - \rho) I_p.
    $$

2.  Compute its Cholesky factor $L$.

3.  For each replication:

    -   Generate a matrix $Z$ with i.i.d. $N(0, 1)$ entries.\
    -   Form the design matrix\
        $$
        X = ZL.
        $$
    -   Draw noise\
        $$
        \epsilon \sim N_n(0, I_n).
        $$
    -   Compute the response\
        $$
        Y = X\beta + \epsilon.
        $$

This procedure is repeated 200 times for each $(\rho, p)$ configuration.

```{r}

set.seed(42)

rho_vals <- c(0.25, 0.50, 0.95)
p_vals   <- c(10, 100, 200)
n        <- 50
rep      <- 200

# Store simulated datasets
results <- list()

for (rho in rho_vals) {
  for (p in p_vals) {

    # Covariance matrix
    Sigma <- rho * matrix(1, p, p) + (1 - rho) * diag(p)
    L <- chol(Sigma)

    sims <- vector("list", rep)

    # True beta
    beta <- rep(0, p)
    beta[1:10] <- 3

    for (r in 1:rep) {

      # Simulate X
      Z <- matrix(rnorm(n * p), n, p)
      X_raw <- Z %*% L

      # Simulate Y = Xβ + ε
      eps <- rnorm(n)
      y <- X_raw %*% beta

      sims[[r]] <- list(X = X_raw, y = as.vector(y + eps))
    }

    results[[paste0("rho_", rho, "_p_", p)]] <- sims
  }
}

```

After generating the simulated datasets, we estimated the regression coefficients using three penalized regression methods: ridge regression, LASSO, and elastic net. All models were fitted using the `glmnet` package, with the tuning parameter $\lambda$ selected via $K$-fold cross-validation. Ridge regression corresponds to $\alpha = 0$, LASSO to $\alpha = 1$, and the elastic net to $\alpha = 0.5$. For LASSO, we considered both the value of $\lambda$ that minimizes the cross-validation error ($\lambda_{min}$) and the more conservative value within one standard error of the minimum ($\lambda_{1se}$).

To speed up computation across the 200 simulated datasets per configuration, the `doParallel` package was used to distribute the repetitions across available CPU cores. For each fitted model we computed the squared estimation error $$
E_\beta=\lVert \hat{\beta} - \beta \rVert_2^2,
$$ and the prediction error $$
E_{X\beta}=\lVert X (\hat{\beta} - \beta) \rVert_2^2.
$$

For each combination of $(\rho, p)$ and each estimation method, we stored the vector of errors across repetitions and their averages. In addition, only the coefficient path, cross-validation curve, and selected $\lambda$ from the final repetition were retained for visualization, rather than saving these objects for every repetition.

```{r, message=FALSE, warning=FALSE, echo=TRUE}

library(glmnet)
library(doParallel)

# Use all available cores
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

results_all <- list()

# Loop over combinations
for (key in names(results)) {

  sims <- results[[key]]
  p <- ncol(sims[[1]]$X)
  beta <- rep(0, p)
  beta[1:10] <- 3

  # Parallel over repetitions
  errors_list <- foreach(r = seq_along(sims), .packages = "glmnet") %dopar% {

    X <- scale(sims[[r]]$X)
    y <- sims[[r]]$y

    res <- list()

    # ---- Ridge (alpha = 0) ----
    ridge_cv <- cv.glmnet(x = X, y = y, alpha = 0)
    ridge_hat <- as.vector(coef(ridge_cv, s = "lambda.min"))[-1]
    res$ridge <- list(
      sq_error   = sum((ridge_hat - beta)^2),
      pred_error = sum((X %*% (ridge_hat - beta))^2),
      cv_fit     = ridge_cv
    )

    # ---- LASSO (alpha = 1) ----
    lasso_cv <- cv.glmnet(x = X, y = y, alpha = 1)

    # lambda.min
    lasso_min_hat <- as.vector(coef(lasso_cv, s = "lambda.min"))[-1]
    res$lasso_min <- list(
      sq_error   = sum((lasso_min_hat - beta)^2),
      pred_error = sum((X %*% (lasso_min_hat - beta))^2),
      cv_fit     = lasso_cv
    )

    # lambda.1se
    lasso_1se_hat <- as.vector(coef(lasso_cv, s = "lambda.1se"))[-1]
    res$lasso_1se <- list(
      sq_error   = sum((lasso_1se_hat - beta)^2),
      pred_error = sum((X %*% (lasso_1se_hat - beta))^2),
      cv_fit     = lasso_cv
    )

    # ---- Elastic Net (alpha = 0.5) ----
    enet_cv <- cv.glmnet(x = X, y = y, alpha = 0.5)
    enet_hat <- as.vector(coef(enet_cv, s = "lambda.min"))[-1]
    res$elastic <- list(
      sq_error   = sum((enet_hat - beta)^2),
      pred_error = sum((X %*% (enet_hat - beta))^2),
      cv_fit     = enet_cv
    )

    res
  }

  # Initialize storage for this combination
  results_all[[key]] <- list()

  # Loop over methods to extract errors, averages, coef paths, and CV curves
  for (method in c("ridge", "lasso_min", "lasso_1se", "elastic")) {

    # Extract errors
    sq_errors <- sapply(errors_list, function(x) x[[method]]$sq_error)
    pred_errors <- sapply(errors_list, function(x) x[[method]]$pred_error)

    # Last repetition's cv_fit
    last_cv_fit <- errors_list[[length(sims)]][[method]]$cv_fit

    # Store lambda selections safely as numeric
    lambda_min <- as.numeric(last_cv_fit$lambda.min)
    lambda_1se <- if(method == "lasso_1se") as.numeric(last_cv_fit$lambda.1se) else NULL

    results_all[[key]][[method]] <- list(
      sq_errors       = sq_errors,
      avg_sq_error    = mean(sq_errors),
      pred_errors     = pred_errors,
      avg_pred_error  = mean(pred_errors),
      coef_path       = coef(last_cv_fit$glmnet.fit),   # single path
      lambda_seq      = last_cv_fit$glmnet.fit$lambda,
      lambda_min      = lambda_min,
      lambda_1se      = lambda_1se,
      cv_curve        = data.frame(
        lambda = last_cv_fit$lambda,
        cvm    = last_cv_fit$cvm,
        cvsd   = last_cv_fit$cvsd
      )
    )
  }

}

stopCluster(cl)


```

```{r}
plot_coef_paths <- function(results_all, method) {
  sparse <- method %in% c("lasso_min", "lasso_1se", "elastic")
  
  n_keys <- length(results_all)
  ncol_plot <- 2
  nrow_plot <- ceiling((n_keys + 1) / ncol_plot)  # +1 for legend slot
  
  # Increase top margin more to make space for numbers and title
  par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 6, 1))
  
  keys <- names(results_all)
  
  for (i in seq_len(nrow_plot * ncol_plot)) {
    if(i <= n_keys){
      key <- keys[i]
      res <- results_all[[key]][[method]]
      
      coef_mat <- as.matrix(res$coef_path)
      if(nrow(coef_mat) > 1){
        coef_mat_no_int <- coef_mat[-1, , drop = FALSE]
      } else {
        coef_mat_no_int <- coef_mat
      }
      
      n_coef <- nrow(coef_mat_no_int)
      ord <- order(res$lambda_seq)
      lambda <- res$lambda_seq[ord]
      coef_mat_no_int <- coef_mat_no_int[, ord, drop = FALSE]
      
      colors <- rep(rgb(0.8, 0.8, 0.8, 0.5), n_coef)
      colors[1:min(10, n_coef)] <- "red"
      
      if(sparse){
        nz_counts <- colSums(coef_mat_no_int != 0)
      }
      
      matplot(
        -log(lambda), t(coef_mat_no_int), type = "l", lty = 1,
        col = colors, xlab = "-log(lambda)", ylab = "Coefficients",
        main = paste("Coefficient Paths:", key, "-", method)
      )
      
      abline(v = -log(res$lambda_min), col = "blue", lwd = 2, lty = 2)
      if(method == "lasso_1se" && !is.null(res$lambda_1se)){
        abline(v = -log(res$lambda_1se), col = "green", lwd = 2, lty = 2)
      }
      
      # Non-zero counts above plot but below title
      if(sparse){
  axis(side = 3, at = -log(lambda), labels = nz_counts, tick = FALSE, line = -0.3, cex.axis = 0.7)
}


      
    } else {
      # Empty plot for legend
      plot.new()
      legend("center",
       legend = if(method == "lasso_1se") c("True beta 1–10", "Zero beta", "lambda.min", "lambda.1se")
                else c("True beta 1–10", "Zero beta", "lambda.min"),
       col = if(method == "lasso_1se") c("red", rgb(0.8,0.8,0.8,0.5), "blue", "green")
                 else c("red", rgb(0.8,0.8,0.8,0.5), "blue"),
       lty = if(method == "lasso_1se") c(1, 1, 2, 2) else c(1, 1, 2),
       cex = 1.5,
       bty = "o",
       box.lwd = 1)
    }
  }
  
  # Reset par
  par(mfrow = c(1,1), mar = c(5, 4, 4, 2) + 0.1)
}








plot_cv_curves <- function(results_all, method) {
  sparse <- method %in% c("lasso_min", "lasso_1se") 
  
  n_keys <- length(results_all)
  ncol_plot <- 2
  nrow_plot <- ceiling((n_keys + 1) / ncol_plot)  # +1 for legend slot
  
  par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 2, 1))
  
  keys <- names(results_all)
  
  for(i in seq_len(nrow_plot * ncol_plot)){
    if(i <= n_keys){
      key <- keys[i]
      res <- results_all[[key]][[method]]
      
      cv_df <- res$cv_curve[order(res$cv_curve$lambda), ]
      lambda <- cv_df$lambda
      
      plot(-log(lambda), cv_df$cvm, type = "l", lwd = 2,
           xlab = "-log(lambda)", ylab = "CV Error",
           main = paste("CV Curve:", key, "-", method))
      
      # ±1 SE lines
      lines(-log(lambda), cv_df$cvm + cv_df$cvsd, lty = 2, col = "gray")
      lines(-log(lambda), cv_df$cvm - cv_df$cvsd, lty = 2, col = "gray")
      
      # Vertical lines for lambda
      abline(v = -log(res$lambda_min), col = "blue", lwd = 2, lty = 2)
      if(method == "lasso_1se" && !is.null(res$lambda_1se)){
        abline(v = -log(res$lambda_1se), col = "green", lwd = 2, lty = 2)
      }
      
    } else {
      # Extra slot for boxed legend
      plot.new()
      legend("center",
             legend = if(sparse) c("CV error", "±1 SE", "lambda.min", "lambda.1se")
                      else c("CV error", "±1 SE", "lambda.min"),
             col = if(sparse) c("black", "gray", "blue", "green") else c("black", "gray", "blue"),
             lty = if(sparse) c(1, 2, 2, 2) else c(1, 2, 2),
             cex = 1.5,
             bty = "o",
             box.lwd = 1)
    }
  }
  
  # Reset par
  par(mfrow = c(1,1), mar = c(5, 4, 4, 2) + 0.1)
}


```

@fig-ridge-cof-paths, bellow, plots the coefficient paths for all the combinations based on one replicate. We observe that the estimates at $p=10$ perform the best which is to be expected given that it corresponds to the true model order. However, we observe that the variance is the lowest for the $\rho=0.95$ replicate, with the high correlation acting as a stabilizing effect when all variables are truly nonzero. However, we see clear issues for the $p>10$ combinations. Due to the correlations present between the zero- and non-zero predictors the non-zero estimates the predictors are all shrunk together, leading to far too low estimates for the non-zero predictors. However, we do see that the non-zero predictors are still somewhat differentiated from those that are zero, as they are mostly consistently shrunk the least.

```{r}
#| label: fig-ridge-cof-paths
#| fig-cap: "Coefficient paths for Ridge Regression"
#| fig-width: 14
#| fig-height: 14
plot_coef_paths(results_all, method = "ridge")

```

@fig-ridge-cv-curve, below, plots the corresponding cross validation curves for the ridge regression. For the $p=10$ combinations, the cross validation errors behave quite similarily for all values of $\rho$ (although the initial error itself is much higher for $\rho=0.95$, with a sigmoid shape. However for higher values of $p$ the error decreases more gradually instead.

```{r}
#| label: fig-ridge-cv-curve
#| fig-cap: "Cross validation curves for Ridge Regression"
#| fig-width: 14
#| fig-height: 14
plot_cv_curves(results_all, method = "ridge")
```

Moving on to LASSO whose coefficient paths are illyustrated in @fig-lasso-cof-paths, below, we see that the estimates at $p=10$ perform similarily to Ridge, although it doesn't gain the same decrease in the spread of the estimates with increased $\rho$. However, for $p>10$ combination there is a large difference. As LASSO can produce sparse estimates we observe that even in the presence of correlation, most of the zero coefficients are set exactly to zero, creating a clear separation from the non-zero coefficients. A few zero coefficients (roughly 15–20 out of 100–200) are not fully suppressed; however, these remaining coefficients are still much lower than the true signals and easily distinguishable in the coefficient paths. The non-zero coefficients themselves are generally slightly below the true value of 3, indicating mild shrinkage, but remain clearly identifiable as signals.

The clear separation between zero and non-zero coefficients begins to weaken only in the most extreme settings — specifically when both the number of predictors and the correlation are very high, where some zero coefficients are less strongly shrunk and the distinction between signal and noise in the coefficient paths is reduced, though the majority of zero coefficients remain far below the non-zero coefficients. However, in the most extreme case of $\rho=0.95$ and $p=200$ two of the non-zero coefficients is incorrectly shrunk to 0.

As for the selection method for $\lambda$, both seems to perform similarily.

```{r}
#| label: fig-lasso-cof-paths
#| fig-cap: "Coefficient paths for LASSO"
#| fig-width: 14
#| fig-height: 22
plot_coef_paths(results_all, method = "lasso_1se")
```

Moving on to the CV error path, iullustrated in @fig-lasso-cv-curve, below, we see that compared to Ridge the behaviour for the cross validation error behaves much similiar across different combinations, with the error first decreasing ands then flattening. However, curiously the $\rho=0.5, p=200$ does not flatten at much. We see that the CV error is similiar for bot values of $\lambda$ except for once again the $\rho=0.5, p=200$ case where it is quite noticibly higher. However, uinless evaluating on test data it is not that meaningful to compare these values.

```{r}
#| label: fig-lasso-cv-curve
#| fig-cap: "Cross validation curves for LASSO"
#| fig-width: 14
#| fig-height: 22
plot_cv_curves(results_all, method = "lasso_1se")
```

We now move on to elastic net, whose coefficient paths are illustrated in @fig-el-cof-paths, below. We see that the performance at the $p=10$ setting is inbetween Ridge and LASSO, with the spread bof the estimates being smaller at for the higfher values of $\rho$ but unable to take fully advantage of the correlation to set the estimates at similiar values. Similarily, elastic net seems to perform somewhere inbeteen Ridge and LASSO for the higehr orders as well. For the values where $\rho$ and/or $p$ are not the most extreme elastic net show a clear separation between zero and non-zero coefficients, with arond the same time of true nulls set to zero. Hower, in the more extreme cases we see that elastic net begins to struggle somewhat compared to LASSO. Separation is slightly worse than that of LASSO in these cases and the non-nulls are shrunk more. Furthermore, in the most extreme case, two non-nulls are incorrectly set to zero while at the same time many more zeroes are kept compared to LASSO.

```{r}
#| label: fig-el-cof-paths
#| fig-cap: "Coefficient paths for Elastic Net"
#| fig-width: 14
#| fig-height: 22
plot_coef_paths(results_all, method = "elastic")
```

The CV error path for the elastic net estimate is illustrated in @fig-el-cv-curve, below. We see that the paths are quite similiar to that of LASSO alhough the $\lambda$ that minimizes CV error tends to be slightly smaller in comparison.

```{r}
#| label: fig-el-cv-curve
#| fig-cap: "Cross validation curves for Elastic net"
#| fig-width: 14
#| fig-height: 22
plot_cv_curves(results_all, method = "elastic")
```

Finally we will look at how the estimationmethods compared overall across all replicates in terms of the two estimation errors $E_\beta$ and $E_{X\beta}$ for each setting. This is illustrated in @tbl-summary, below. We observe that in the $p=10$, setting Ridge performs the best for both metriucs, , the one diffewrent is when $rho=0.95$ is for $\rho=0.95$, where elastic net performs better in the prediction error only. In higher dimensions LASSO with $\lambda_{\min}$ performs the best in all cases. This was clearly visible in the graphs and given that we have not evaluated our model on unseen data it is naturally that \$\\lambda\_{\\min} should outperform $\lambda_{1\text{se}}$ as well.

```{r}
#| echo: false
#| label: tbl-summary
#| fig-cap: "Summary of Estimation and Prediction Errors with best method bolded"
library(dplyr)
library(knitr)
library(kableExtra)

methods <- c("ridge", "lasso_min", "lasso_1se", "elastic")

big_table <- bind_rows(lapply(names(results_all), function(key) {
  df <- t(sapply(methods, function(m) c(
    E_beta = mean(results_all[[key]][[m]]$sq_error),
    E_Xbeta = mean(results_all[[key]][[m]]$pred_error)
  )))
  df <- as.data.frame(df, stringsAsFactors = FALSE)
  df$Method <- rownames(df)
  df$Setting <- key
  rownames(df) <- NULL
  df
}))

big_table <- big_table %>%
  mutate(across(c(E_beta, E_Xbeta), as.numeric)) %>%
  group_by(Setting) %>%
  mutate(
    E_beta = ifelse(E_beta == min(E_beta), paste0("\\textbf{", round(E_beta,3), "}"), round(E_beta,3)),
    E_Xbeta = ifelse(E_Xbeta == min(E_Xbeta), paste0("\\textbf{", round(E_Xbeta,3), "}"), round(E_Xbeta,3))
  ) %>%
  ungroup() %>%
  select(Setting, Method, E_beta, E_Xbeta)

kable(big_table, digits = 3, booktabs = TRUE, escape = FALSE)


```

This result can at first look seem a bit unexpected given that Elastic Net is generally attuned to settings with correlated variables. However, in this experiment, all variables are equally correlated, which diminishes the typical advantage of Elastic Net in selecting groups of predictors. LASSO tends to select only a few predictors and set the rest to zero, which in this uniform correlation setting still preserves most true signals, explaining why it performs surprisingly well. Ridge, on the other hand, shrinks all coefficients more evenly, which helps in low-dimensional non-sparse settings by stabilizing estimates but provides less selective power in high dimensions. Elastic Net behaves similarly to LASSO here but does not outperform it because the uniform correlation prevents it from favoring grouped variables.

# Part 2: Diabetes data

We now move on to the next part where we will work with real data, specifically we will attempt to model disease prediction of diabetes,. The data have been standaridez meaning that wee can directly compare effect of variables in terms of magniotuyde much nmor easiy.

```{r}
#|echo: false
diabetes <- read.csv("diabetesdata.csv")
```

We will begn by taking a short look at the correlations in the dataset. We see that there are many moderate to strong correlations present between the blood serum measurements indicating that there might be multicollinearity present.

```{r}
#| echo: false
#| label: fig-diab-cor
#| fig-cap: "Correlation plot of the Diabetes dataset"
library(corrplot)
cor_matrix <- cor(diabetes)
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45)
```

We begin by fitting an orinary leasr squares regression using all the predictors. The estimated model is described in @tbl-ols-mod, below. We see that of the non blood serum measurements, all but *age* are significant. However, among the blood serum measurements only S5 is significant.

```{r}
#|echo: false
#|label: tbl-ols-mod
model <- lm(Y ~ ., data = diabetes)

# Extract coefficients
coef_table <- summary(model)$coefficients

# Display coefficients as a kable
kable(coef_table, digits = 4, caption = "OLS Regression Coefficients")

```

@tbl-ols-metrics, below, illustrates summary metrics for the OLS model. With $R^2$ around 0.5 our model can be said to explain a moderate amount of disease progression for diabetes.

```{r}
#|echo: false
#|label: tbl-ols-metrics
#|# Extract key model statistics
model_stats <- data.frame(
  Statistic = c("Residual Standard Error", "Adjusted R-squared", "Multiple R-squared", "F-statistic"),
  Value = c(summary(model)$sigma,
            summary(model)$adj.r.squared,
            summary(model)$r.squared,
            summary(model)$fstatistic[1])
)

# Display model summary statistics as a kable
kable(model_stats, digits = 4, caption = "OLS Model Summary Statistics")
```

We now move on to attempting to fit a model using regularized methods, beginning with ridge regression.

```{r}
set.seed(42)
#| echo: false
# Load necessary libraries
library(ggplot2)
library(reshape2)

# Load data
diabetes <- read.csv("diabetesdata.csv")

# Separate predictors and response
X <- as.matrix(diabetes[, -which(names(diabetes) == "Y")])
y <- diabetes$Y


# Cross-validation to select lambda
set.seed(42)
cv_ridge <- cv.glmnet(X, y, alpha = 0, standardize = FALSE)

# Use the same lambda grid as in cv.glmnet to fit the model
ridge_model <- glmnet(X, y, alpha = 0, standardize = FALSE, lambda = cv_ridge$glmnet.fit$lambda)
```

The below output illustrates the Ridge Regression Coefficents at $\lambda_{\min}$. We see that the bloood serum measurments (along with sex which was correlated to S3) have been shrunk to have more similiar level, compared to before when a few where much larger.

```{r}
coef(ridge_model, s = cv_ridge$lambda.min)
```

@fig-dcof-ridge, below, illustrates the coefficient path for the ridge regression. We see that the blood serum variables move together and that *sex* is shrunk quite aggresivly.

```{r}
#| echo: false
#| label: fig-dcof-ridge
#| fig-cap: "Coefficient path for Ridge regression on Diabetes dataset."

coef_mat <- as.matrix(ridge_model$beta)
coef_df <- data.frame(lambda = ridge_model$lambda, t(coef_mat))
coef_melt <- melt(coef_df, id.vars = "lambda")

ggplot(coef_melt, aes(x = -log(lambda), y = value, color = variable)) +
  geom_line(size = 1) +
  geom_vline(xintercept = -log(cv_ridge$lambda.min), linetype = "dashed", color = "red") +
  labs(x = "-log(lambda)", y = "Coefficient",
       title = "Ridge Regression Coefficient Paths",
       subtitle = "Red dashed line indicates lambda.min") +
  theme_minimal() +
  theme(legend.title = element_blank())

```

s @fig-dcv-ridge, below, illustrates the cross validation error path for the ridge regressio. We see that the error decreases smoothly rather than having a sharp plateau.

```{r}
#| echo: false
#| label: fig-dcv-ridge
#| fig-cap: "Cross validation error path for Ridge regression on Diabetes dataset."
# Prepare CV data
cv_df <- data.frame(
  lambda = cv_ridge$lambda,
  cvm = cv_ridge$cvm,   # mean CV error
  cvup = cv_ridge$cvup, # upper SE
  cvlo = cv_ridge$cvlo  # lower SE
)

# Plot CV curve with -log(lambda)
ggplot(cv_df, aes(x = -log(lambda), y = cvm)) +
  geom_line(size = 1, color = "blue") +
  geom_ribbon(aes(ymin = cvlo, ymax = cvup), alpha = 0.2) +
  geom_vline(xintercept = -log(cv_ridge$lambda.min), linetype = "dashed", color = "red") +
  labs(
    x = "-log(lambda)",
    y = "Mean CV Error",
    title = "Cross-Validation Curve for Ridge Regression",
    subtitle = "Red dashed = lambda.min"
  ) +
  theme_minimal()


```

```{r}
#| echo: false
lasso_mod <- glmnet(X, y, alpha = 1, standardize = FALSE)


set.seed(42)
cv_lasso_mod <- cv.glmnet(X, y, alpha = 1, standardize = FALSE)

# Best lambda (lambda.min)
lambda_best <- cv_lasso_mod$lambda.min


coef_lasso <- coef(cv_lasso_mod, s = "lambda.min")

# Convert to data frame for reporting
coef_lasso_df <- data.frame(
  Variable = rownames(coef_lasso),
  Coefficient = as.numeric(coef_lasso)
)



```

The blow output illustrates the estimated coefficients for the LASSO coefficents. While some coefficents have been somewhat shrunk the estimates are overall quite similiar to the OLS model and no coefficents have been set to 0.

```{r}
coef(lasso_mod, s = cv_lasso_mod$lambda.min)
```

s @fig-dcof-lasso, below, illustrates the coefficent paths for LASSO. We see that at higher $\lambda$ many of the estimates are set to close to 0, but that S5 and *Sex* start to quickly approach their OLS estimates once they enter the model.

```{r}
#| echo: false
#| label: fig-dcof-lasso
#| fig-cap: "Coefficient path for LASSO on Diabetes dataset."


# Extract coefficients from Lasso model
coef_mat <- as.matrix(lasso_mod$beta)
coef_df <- data.frame(lambda = lasso_mod$lambda, t(coef_mat))
coef_melt <- melt(coef_df, id.vars = "lambda")

# Plot with -log(lambda)
ggplot(coef_melt, aes(x = -log(lambda), y = value, color = variable)) +
  geom_line(size = 1) +
  geom_vline(xintercept = -log(cv_lasso_mod$lambda.min), linetype = "dashed", color = "red") +
  labs(x = "-log(lambda)", y = "Coefficient",
       title = "Lasso Regression Coefficient Paths",
       subtitle = "Red dashed line indicates lambda.min") +
  theme_minimal() +
  theme(legend.title = element_blank())


```

 s @fig-dcv-lasso, below, illustrates the cross validation error path for LASSO. Unlike Ridge the path has a large plateau and it seems that using $\lambda_{1\text{se}}$ instead and testing the model on unseen data could potentionally be fruitful, given how different the model looks at the start of the pleateau compared to the end in the coefficent paths.

```{r}
#| echo: false
#| label: fig-dcv-lasso
#| fig-cap: "Cross validation error path for LASSO on Diabetes dataset."
# Prepare CV data
cv_df <- data.frame(
  lambda = cv_lasso_mod$lambda,
  cvm = cv_lasso_mod$cvm,   # mean CV error
  cvup = cv_lasso_mod$cvup, # upper SE
  cvlo = cv_lasso_mod$cvlo  # lower SE
)

# Plot CV curve with -log(lambda)
ggplot(cv_df, aes(x = -log(lambda), y = cvm)) +
  geom_line(size = 1, color = "blue") +
  geom_ribbon(aes(ymin = cvlo, ymax = cvup), alpha = 0.2) +
  geom_vline(xintercept = -log(cv_lasso_mod$lambda.min), linetype = "dashed", color = "red") +
  labs(
    x = "-log(lambda)",
    y = "Mean CV Error",
    title = "Cross-Validation Curve for Lasso Regression",
    subtitle = "Red dashed = lambda.min"
  ) +
  theme_minimal()

```

In order to evaluate the performance of the models, we performed 10-fold cross-validation (CV). In this procedure, the dataset is randomly split into 10 equally sized subsets, or “folds.” For each iteration, one fold is held out as a test set while the model is trained on the remaining nine folds. The trained model is then used to predict the responses for the held-out fold, and the mean squared error (MSE) is calculated. This process is repeated for all 10 folds, and the resulting MSE values are averaged to provide an estimate of the model’s predictive performance on unseen data.

For OLS, the model was directly fitted on the training folds in each iteration. For Ridge and Lasso, the cross-validation procedure used the previously selected optimal regularization parameter, $\lambda_{\min}$,⁡ obtained from CV on the full dataset. Since the best lambda was chosen using information from the entire dataset, this introduces a slight “data peeking”, meaning that the reported CV errors for Ridge and Lasso may be slightly optimistic.

For Lasso, it could also have been interesting to evaluate performance using $\lambda_{1\text{se}}$, which provides a more conservative choice that often results in sparser models. However, we did not use it in this analysis because it was not specified. Despite this, the procedure still provides a reasonable comparison of predictive performance across models while accounting for variability due to different training-test splits.

The resulting MSE values are decribed in @tbl-10fold-mse, below, We observe that OLS performed the best with LASSO performing just slightlly better (completely natural, given that the shrinkage was very low). Ridge performs slightly worse.

```{r}
#|echo: false
#|label: tbl-10fold-mse
#|tbl-cap: "MSE from 10-fold CV"
library(caret)

set.seed(42)

# 1. Prepare data
X <- as.matrix(diabetes[, -which(names(diabetes) == "Y")])
y <- diabetes$Y

# 2. Define 10-fold CV
folds <- createFolds(y, k = 10, list = TRUE, returnTrain = TRUE)

# -------------------------
# 3. OLS CV
# -------------------------
ols_mse_vec <- c()

for (i in 1:10) {
  train_idx <- folds[[i]]
  test_idx <- setdiff(1:nrow(X), train_idx)
  
  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  
  X_test <- X[test_idx, ]
  y_test <- y[test_idx]
  
  # Proper data frames for lm
  data_train <- data.frame(y = y_train, X_train)
  data_test  <- data.frame(y = y_test, X_test)
  
  ols_fit <- lm(y ~ ., data = data_train)
  y_pred <- predict(ols_fit, newdata = data_test)
  
  ols_mse_vec <- c(ols_mse_vec, mean((y_test - y_pred)^2))
}

ols_mse <- mean(ols_mse_vec)

# -------------------------
# 4. Ridge CV (alpha = 0)
# -------------------------
cv_ridge <- cv.glmnet(X, y, alpha = 0, standardize = FALSE)
ridge_best_lambda <- cv_ridge$lambda.min

ridge_mse_vec <- c()
for (i in 1:10) {
  train_idx <- folds[[i]]
  test_idx <- setdiff(1:nrow(X), train_idx)
  
  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  
  X_test <- X[test_idx, ]
  y_test <- y[test_idx]
  
  ridge_fit <- glmnet(X_train, y_train, alpha = 0, lambda = ridge_best_lambda, standardize = FALSE)
  y_pred <- predict(ridge_fit, newx = X_test)
  
  ridge_mse_vec <- c(ridge_mse_vec, mean((y_test - y_pred)^2))
}

ridge_mse <- mean(ridge_mse_vec)

# -------------------------
# 5. Lasso CV (alpha = 1)
# -------------------------
cv_lasso <- cv.glmnet(X, y, alpha = 1, standardize = FALSE)
lasso_best_lambda <- cv_lasso$lambda.min

lasso_mse_vec <- c()
for (i in 1:10) {
  train_idx <- folds[[i]]
  test_idx <- setdiff(1:nrow(X), train_idx)
  
  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  
  X_test <- X[test_idx, ]
  y_test <- y[test_idx]
  
  lasso_fit <- glmnet(X_train, y_train, alpha = 1, lambda = lasso_best_lambda, standardize = FALSE)
  y_pred <- predict(lasso_fit, newx = X_test)
  
  lasso_mse_vec <- c(lasso_mse_vec, mean((y_test - y_pred)^2))
}

lasso_mse <- mean(lasso_mse_vec)

# -------------------------
# 6. Combine results
# -------------------------
mse_comparison <- data.frame(
  Model = c("OLS", "Ridge", "Lasso"),
  CV_MSE = c(ols_mse, ridge_mse, lasso_mse)
)

kable(mse_comparison, digits = 2)


```

In terms of coefficient shrinkage, OLS does not apply any regularization. This can result in large differences in magnitude between predictors, particularly when some are correlated or have higher variance. Ridge regression, on the other hand, shrinks all coefficients towards zero, though none are set exactly to zero. In this dataset, the shrinkage is modest, but Ridge reduces extreme coefficient values, especially for correlated blood serum measurements (and *Sex*), providing more stable estimates. Lasso performs both shrinkage and variable selection, potentially setting some coefficients exactly to zero. However, at $\lambda_{\min}$ the shrinkage was small and no coefficients were removed, so the estimates remain quite similar to OLS. However, using a more conservative choice such as $\lambda_{1\text{se}}$ could perhaps produce a sparser, more interpretable model. On the other hand, eliminating correlated variables could make the model simpler, it would also discard useful information. Ideally, we would prefer to retain all informative predictors while mitigating the negative effects of multicollinearity through shrinkage or other methods.

Regarding model interpretability, OLS is straightforward, as all variables are retained and coefficients can be directly compared. Ridge is slightly less interpretable because coefficients are shrunk, but it does not perform variable selection, so all predictors remain in the model. Lasso is potentially the most interpretable, as it can highlight the most important predictors by eliminating uninformative ones, but in this analysis it was overall very similiar to OLS.

Given that the regularized methods provide only modest gains and no variables were actually eliminated, OLS might be considered the best choice for this dataset, producing unbiased estimates while retaining all informative predictors.
