---
title: "Computer Lab 4: High Dimensional Data Analysis"
author: "Sara Kopelman"
date: today
date-format: long
format:
  pdf:
    toc: false
    number-sections: false
    booktabs: true
header-includes:
  - \usepackage{float}  
---

# Part 1: Monte Carlo simulation

In this section, we simulate data from the linear model

$$
Y = X\beta + \epsilon
$$

where $X$ is an $n \times p$ design matrix drawn from a multivariate normal distribution with independent rows

$$
X_i \sim N_p(0, \Sigma)
$$

and covariance matrix

$$
\Sigma = \rho J_p + (1 - \rho) I_p,
$$

where $J_p$ is the $p \times p$ matrix of ones, $I_p$ is the $p \times p$ identity matrix, and $\rho$ denotes the pairwise correlation among covariates.

The true coefficient vector is given by

$$
\beta_1 = \beta_2 = \cdots = \beta_{10} = 3,
\qquad
\beta_{11} = \cdots = \beta_p = 0,
$$

and the noise term satisfies

$$
\epsilon \sim N_n(0, I_n).
$$

We consider all combinations of

$$
\rho \in \{0.25, 0.50, 0.95\},
\qquad
p \in \{10, 100, 200\},
$$

with a fixed sample size $n = 50$. For each configuration, 200 independent datasets are generated.

For each pair $(\rho, p)$, the simulation steps are:

1.  Construct the covariance matrix\
    $$
    \Sigma = \rho J_p + (1 - \rho) I_p.
    $$

2.  Compute its Cholesky factor $L$.

3.  For each replication:

    -   Generate a matrix $Z$ with i.i.d. $N(0, 1)$ entries.\
    -   Form the design matrix\
        $$
        X = ZL.
        $$
    -   Draw noise\
        $$
        \epsilon \sim N_n(0, I_n).
        $$
    -   Compute the response\
        $$
        Y = X\beta + \epsilon.
        $$

This procedure is repeated 200 times for each $(\rho, p)$ configuration. The exact procedure is described in the below code chunk.

```{r}

set.seed(42)

rho_vals <- c(0.25, 0.50, 0.95)
p_vals   <- c(10, 100, 200)
n        <- 50
rep      <- 200

# Store simulated datasets
results <- list()

for (rho in rho_vals) {
  for (p in p_vals) {

    # Covariance matrix
    Sigma <- rho * matrix(1, p, p) + (1 - rho) * diag(p)
    L <- chol(Sigma)

    sims <- vector("list", rep)

    # True beta
    beta <- rep(0, p)
    beta[1:10] <- 3

    for (r in 1:rep) {

      # Simulate X
      Z <- matrix(rnorm(n * p), n, p)
      X_raw <- Z %*% L

      # Simulate Y = Xβ + ε
      eps <- rnorm(n)
      y <- X_raw %*% beta

      sims[[r]] <- list(X = X_raw, y = as.vector(y + eps))
    }

    results[[paste0("rho_", rho, "_p_", p)]] <- sims
  }
}

```

After generating the simulated data-sets, we estimated the regression coefficients using three regularized methods: ridge regression, LASSO, and elastic net. All models were fitted using the `glmnet` package, with the tuning parameter $\lambda$ selected via $K$-fold cross-validation. Ridge regression corresponds to $\alpha = 0$, LASSO to $\alpha = 1$, and the elastic net to $\alpha = 0.5$. For LASSO, we considered both the value of $\lambda$ that minimizes the cross-validation error ($\lambda_{min}$) and the more conservative value within one standard error of the minimum ($\lambda_{1se}$).

To speed up computation across the 200 simulated datasets per configuration, the `doParallel` package was used to distribute the repetitions across available CPU cores. For each fitted model we computed the squared estimation error $$
E_\beta=\lVert \hat{\beta} - \beta \rVert_2^2,
$$ and the prediction error $$
E_{X\beta}=\lVert X (\hat{\beta} - \beta) \rVert_2^2.
$$

For each combination of $(\rho, p)$ and each estimation method, we stored the vector of errors across repetitions and their averages. In addition, only the coefficient path, cross-validation curve, and selected $\lambda$ from the final repetition were retained for visualization, rather than saving these objects for every repetition. The exact estimation procedure is described in the below code chunk.

```{r, message=FALSE, warning=FALSE, echo=TRUE}

library(glmnet)
library(doParallel)

# Use all available cores
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

results_all <- list()

# Loop over combinations
for (key in names(results)) {

  sims <- results[[key]]
  p <- ncol(sims[[1]]$X)
  beta <- rep(0, p)
  beta[1:10] <- 3

  # Parallel over repetitions
  errors_list <- foreach(r = seq_along(sims), .packages = "glmnet") %dopar% {

    X <- scale(sims[[r]]$X)
    y <- sims[[r]]$y

    res <- list()

    # ---- Ridge (alpha = 0) ----
    ridge_cv <- cv.glmnet(x = X, y = y, alpha = 0)
    ridge_hat <- as.vector(coef(ridge_cv, s = "lambda.min"))[-1]
    res$ridge <- list(
      sq_error   = sum((ridge_hat - beta)^2),
      pred_error = sum((X %*% (ridge_hat - beta))^2),
      cv_fit     = ridge_cv
    )

    # ---- LASSO (alpha = 1) ----
    lasso_cv <- cv.glmnet(x = X, y = y, alpha = 1)

    # lambda.min
    lasso_min_hat <- as.vector(coef(lasso_cv, s = "lambda.min"))[-1]
    res$lasso_min <- list(
      sq_error   = sum((lasso_min_hat - beta)^2),
      pred_error = sum((X %*% (lasso_min_hat - beta))^2),
      cv_fit     = lasso_cv
    )

    # lambda.1se
    lasso_1se_hat <- as.vector(coef(lasso_cv, s = "lambda.1se"))[-1]
    res$lasso_1se <- list(
      sq_error   = sum((lasso_1se_hat - beta)^2),
      pred_error = sum((X %*% (lasso_1se_hat - beta))^2),
      cv_fit     = lasso_cv
    )

    # ---- Elastic Net (alpha = 0.5) ----
    enet_cv <- cv.glmnet(x = X, y = y, alpha = 0.5)
    enet_hat <- as.vector(coef(enet_cv, s = "lambda.min"))[-1]
    res$elastic <- list(
      sq_error   = sum((enet_hat - beta)^2),
      pred_error = sum((X %*% (enet_hat - beta))^2),
      cv_fit     = enet_cv
    )

    res
  }

  # Initialize storage for this combination
  results_all[[key]] <- list()

  # Loop over methods to extract errors, averages, coef paths, and CV curves
  for (method in c("ridge", "lasso_min", "lasso_1se", "elastic")) {

    # Extract errors
    sq_errors <- sapply(errors_list, function(x) x[[method]]$sq_error)
    pred_errors <- sapply(errors_list, function(x) x[[method]]$pred_error)

    # Last repetition's cv_fit
    last_cv_fit <- errors_list[[length(sims)]][[method]]$cv_fit

    # Store lambda selections safely as numeric
    lambda_min <- as.numeric(last_cv_fit$lambda.min)
    lambda_1se <- if(method == "lasso_1se") as.numeric(last_cv_fit$lambda.1se) else NULL

    results_all[[key]][[method]] <- list(
      sq_errors       = sq_errors,
      avg_sq_error    = mean(sq_errors),
      pred_errors     = pred_errors,
      avg_pred_error  = mean(pred_errors),
      coef_path       = coef(last_cv_fit$glmnet.fit),   # single path
      lambda_seq      = last_cv_fit$glmnet.fit$lambda,
      lambda_min      = lambda_min,
      lambda_1se      = lambda_1se,
      cv_curve        = data.frame(
        lambda = last_cv_fit$lambda,
        cvm    = last_cv_fit$cvm,
        cvsd   = last_cv_fit$cvsd
      )
    )
  }

}

stopCluster(cl)


```

```{r}
#| echo: false
plot_coef_paths <- function(results_all, method) {
  sparse <- method %in% c("lasso_min", "lasso_1se", "elastic")
  
  n_keys <- length(results_all)
  ncol_plot <- 2
  nrow_plot <- ceiling((n_keys + 1) / ncol_plot)  # +1 for legend slot
  
  # Increase top margin more to make space for numbers and title
  par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 6, 1))
  
  keys <- names(results_all)
  
  for (i in seq_len(nrow_plot * ncol_plot)) {
    if(i <= n_keys){
      key <- keys[i]
      res <- results_all[[key]][[method]]
      
      coef_mat <- as.matrix(res$coef_path)
      if(nrow(coef_mat) > 1){
        coef_mat_no_int <- coef_mat[-1, , drop = FALSE]
      } else {
        coef_mat_no_int <- coef_mat
      }
      
      n_coef <- nrow(coef_mat_no_int)
      ord <- order(res$lambda_seq)
      lambda <- res$lambda_seq[ord]
      coef_mat_no_int <- coef_mat_no_int[, ord, drop = FALSE]
      
      colors <- rep(rgb(0.8, 0.8, 0.8, 0.5), n_coef)
      colors[1:min(10, n_coef)] <- "red"
      
      if(sparse){
        nz_counts <- colSums(coef_mat_no_int != 0)
      }
      
      matplot(
        -log(lambda), t(coef_mat_no_int), type = "l", lty = 1,
        col = colors, xlab = "-log(lambda)", ylab = "Coefficients",
        main = paste("Coefficient Paths:", key, "-", method)
      )
      
      abline(v = -log(res$lambda_min), col = "blue", lwd = 2, lty = 2)
      if(method == "lasso_1se" && !is.null(res$lambda_1se)){
        abline(v = -log(res$lambda_1se), col = "green", lwd = 2, lty = 2)
      }
      
      # Non-zero counts above plot but below title
      if(sparse){
  axis(side = 3, at = -log(lambda), labels = nz_counts, tick = FALSE, line = -0.3, cex.axis = 0.7)
}


      
    } else {
      # Empty plot for legend
      plot.new()
      legend("center",
       legend = if(method == "lasso_1se") c("True beta 1–10", "Zero beta", "lambda.min", "lambda.1se")
                else c("True beta 1–10", "Zero beta", "lambda.min"),
       col = if(method == "lasso_1se") c("red", rgb(0.8,0.8,0.8,0.5), "blue", "green")
                 else c("red", rgb(0.8,0.8,0.8,0.5), "blue"),
       lty = if(method == "lasso_1se") c(1, 1, 2, 2) else c(1, 1, 2),
       cex = 1.5,
       bty = "o",
       box.lwd = 1)
    }
  }
  
  # Reset par
  par(mfrow = c(1,1), mar = c(5, 4, 4, 2) + 0.1)
}








plot_cv_curves <- function(results_all, method) {
  sparse <- method %in% c("lasso_min", "lasso_1se") 
  
  n_keys <- length(results_all)
  ncol_plot <- 2
  nrow_plot <- ceiling((n_keys + 1) / ncol_plot)  # +1 for legend slot
  
  par(mfrow = c(nrow_plot, ncol_plot), mar = c(4, 4, 2, 1))
  
  keys <- names(results_all)
  
  for(i in seq_len(nrow_plot * ncol_plot)){
    if(i <= n_keys){
      key <- keys[i]
      res <- results_all[[key]][[method]]
      
      cv_df <- res$cv_curve[order(res$cv_curve$lambda), ]
      lambda <- cv_df$lambda
      
      plot(-log(lambda), cv_df$cvm, type = "l", lwd = 2,
           xlab = "-log(lambda)", ylab = "CV Error",
           main = paste("CV Curve:", key, "-", method))
      
      # ±1 SE lines
      lines(-log(lambda), cv_df$cvm + cv_df$cvsd, lty = 2, col = "gray")
      lines(-log(lambda), cv_df$cvm - cv_df$cvsd, lty = 2, col = "gray")
      
      # Vertical lines for lambda
      abline(v = -log(res$lambda_min), col = "blue", lwd = 2, lty = 2)
      if(method == "lasso_1se" && !is.null(res$lambda_1se)){
        abline(v = -log(res$lambda_1se), col = "green", lwd = 2, lty = 2)
      }
      
    } else {
      # Extra slot for boxed legend
      plot.new()
      legend("center",
             legend = if(sparse) c("CV error", "±1 SE", "lambda.min", "lambda.1se")
                      else c("CV error", "±1 SE", "lambda.min"),
             col = if(sparse) c("black", "gray", "blue", "green") else c("black", "gray", "blue"),
             lty = if(sparse) c(1, 2, 2, 2) else c(1, 2, 2),
             cex = 1.5,
             bty = "o",
             box.lwd = 1)
    }
  }
  
  # Reset par
  par(mfrow = c(1,1), mar = c(5, 4, 4, 2) + 0.1)
}


```

We begin by looking at Ridge regression. @fig-ridge-cof-paths, below, plots the coefficient paths for Ridge across all combinations based on one replicate. We observe that the estimates at $p=10$ perform the best, as expected, since this corresponds to the true model order. The variance is lowest for the $\rho=0.95$ replicate, with high correlation stabilizing the estimates when all variables are truly nonzero. Clear issues arise for the $p>10$ combinations: due to correlations between zero and non-zero predictors, the non-zero estimates are all shrunk together, resulting in substantially underestimated values for the true signals. Nevertheless, the non-zero predictors remain somewhat distinguishable from the zeros, as they are generally shrunk the least.

```{r}
#| echo: false
#| label: fig-ridge-cof-paths
#| fig-cap: "Coefficient paths for Ridge Regression"
#| fig-width: 14
#| fig-height: 22
#| fig-pos: "H"
plot_coef_paths(results_all, method = "ridge")

```

@fig-ridge-cv-curve, below, plots the corresponding cross validation curves for the ridge regression. For the $p=10$ combinations, the cross validation errors behave quite similarly for all values of $\rho$ (although the initial error itself is much higher for $\rho=0.95$, with a sigmoid shape). However, for higher values of $p$ the error decreases more gradually instead.

```{r}
#| label: fig-ridge-cv-curve
#| fig-cap: "Cross validation curves for Ridge Regression"
#| fig-width: 14
#| fig-height: 22
#| echo: false
#| fig-pos: "H"
plot_cv_curves(results_all, method = "ridge")
```

Moving on to LASSO, whose coefficient paths are illustrated in @fig-lasso-cof-paths, we see that the estimates at $p=10$ perform similarly to Ridge, although LASSO does not achieve the same reduction in the spread of estimates with increasing $\rho$. For $p>10$ combinations, however, differences become more pronounced. Because LASSO can produce sparse estimates, even in the presence of correlation, most of the zero coefficients are set exactly to zero, creating a clear separation from the non-zero coefficients. A few zero coefficients (roughly 15–30 out of 100–200) are not fully suppressed, but these remaining coefficients are still much lower than the true signals and easily distinguishable in the coefficient paths. The non-zero coefficients themselves are generally slightly below the true value of 3, indicating mild shrinkage, but they remain clearly identifiable as signals.

The clear separation between zero and non-zero coefficients begins to weaken only in the most extreme settings, specifically when both the number of predictors and the correlation are very high. In these cases, some zero coefficients are less strongly shrunk, reducing the distinction between signal and noise in the coefficient paths, though the majority of zero coefficients remain far below the non-zero coefficients. In the most extreme case of $\rho=0.95$ and $p=200$, one of the non-zero coefficients are incorrectly shrunk to zero.

As for the selection method for $\lambda$, both seems to perform similarly in terms of the estimates and shrinkage.

```{r}
#| label: fig-lasso-cof-paths
#| fig-cap: "Coefficient paths for LASSO"
#| fig-width: 14
#| fig-height: 22
#| echo: false
#| fig-pos: "H"
plot_coef_paths(results_all, method = "lasso_1se")
```

Moving on to the CV error path, illustrated in @fig-lasso-cv-curve, we see that, compared to Ridge, the cross-validation error behaves very similarly across different combinations: it first decreases and then flattens. However, curiously, the $\rho=0.5, p=200$ case does not flatten as much. For most combinations, the CV error is similar across values of $\lambda$, except again for the $\rho=0.5, p=200$ case, where it is noticeably higher. That said, unless the error is evaluated on independent test data, comparing these absolute values is not particularly meaningful.

```{r}
#| label: fig-lasso-cv-curve
#| fig-cap: "Cross validation curves for LASSO"
#| fig-width: 14
#| fig-height: 22
#| echo: false
#| fig-pos: "H"
plot_cv_curves(results_all, method = "lasso_1se")
```

We now move on to Elastic Net, whose coefficient paths are illustrated in @fig-el-cof-paths, below. We see that performance at the $p=10$ setting lies between Ridge and LASSO: the spread of the estimates is smaller for higher values of $\rho$, but Elastic Net is unable to fully utilize the correlation to align the estimates. Similarly, for higher orders, Elastic Net performs between Ridge and LASSO. In settings where $\rho$ and/or $p$ are not extreme, Elastic Net shows a clear separation between zero and non-zero coefficients, with roughly the same proportion of true nulls set to zero. However, in more extreme cases, Elastic Net begins to struggle compared to LASSO. Separation between zero and non-zero coefficients is slightly worse, the non-zero coefficients are shrunk more, and in the most extreme case, two non-zero coefficients are incorrectly set to zero while at the same time many more zero coefficients are retained compared to LASSO.

```{r}
#| label: fig-el-cof-paths
#| fig-cap: "Coefficient paths for Elastic Net"
#| fig-width: 14
#| fig-height: 22
#| echo: false
#| fig-pos: "H"
plot_coef_paths(results_all, method = "elastic")
```

The CV error path for the Elastic Net estimates is illustrated in @fig-el-cv-curve, below. We see that the paths are quite similar to those of LASSO, although the value of $\lambda$ that minimizes the CV error tends to be slightly smaller.

```{r}
#| label: fig-el-cv-curve
#| fig-cap: "Cross validation curves for Elastic net"
#| fig-width: 14
#| fig-height: 22
#| echo: false
#| fig-pos: "H"
plot_cv_curves(results_all, method = "elastic")
```

Finally, we examine how the estimation methods compare overall across all replicates in terms of the two estimation errors, $E_\beta$ and $E_{X\beta}$, for each setting. This is illustrated in @tbl-summary, below. We observe that in the $p=10$ setting, Ridge performs the best for both metrics, with the only exception being for $\rho=0.95$, where Elastic Net achieves a slightly better prediction error. In higher dimensions, LASSO with $\lambda_{\min}$ performs the best in all cases. This is consistent with the coefficient path graphs, and given that we have not evaluated the models on unseen data, it is natural that $\lambda_{\min}$ outperforms $\lambda_{1\text{se}}$ as well

```{r, message=FALSE, warning=FALSE}
#| echo: false
#| label: tbl-summary
#| tbl-cap: "Summary of Estimation and Prediction Errors with best method bolded"
library(dplyr)
library(knitr)
library(kableExtra)

methods <- c("ridge", "lasso_min", "lasso_1se", "elastic")

big_table <- bind_rows(lapply(names(results_all), function(key) {
  df <- t(sapply(methods, function(m) c(
    E_beta = mean(results_all[[key]][[m]]$sq_error),
    E_Xbeta = mean(results_all[[key]][[m]]$pred_error)
  )))
  df <- as.data.frame(df, stringsAsFactors = FALSE)
  df$Method <- rownames(df)
  df$Setting <- key
  rownames(df) <- NULL
  df
}))

big_table <- big_table %>%
  mutate(across(c(E_beta, E_Xbeta), as.numeric)) %>%
  group_by(Setting) %>%
  mutate(
    E_beta = ifelse(E_beta == min(E_beta), paste0("\\textbf{", round(E_beta,3), "}"), round(E_beta,3)),
    E_Xbeta = ifelse(E_Xbeta == min(E_Xbeta), paste0("\\textbf{", round(E_Xbeta,3), "}"), round(E_Xbeta,3))
  ) %>%
  ungroup() %>%
  select(Setting, Method, E_beta, E_Xbeta)

kable(big_table, digits = 3, booktabs = TRUE, escape = FALSE)


```

This result can at first look seem a bit unexpected given that Elastic Net is generally attuned to settings with correlated variables. However, in this experiment, all variables are equally correlated, which diminishes the typical advantage of Elastic Net in selecting groups of predictors. LASSO tends to select only a few predictors and set the rest to zero, which in this uniform correlation setting still preserves most true signals, explaining why it performs surprisingly well. Ridge, on the other hand, shrinks all coefficients more evenly, which helps in low-dimensional non-sparse settings by stabilizing estimates but provides less selective power in high dimensions. Elastic Net behaves similarly to LASSO here but does not outperform it because the uniform correlation prevents it from favoring grouped variables.

# Part 2: Diabetes data

We now move on to the next part, where we will work with real data. Specifically, we will attempt to model disease prediction for diabetes. The data have been standardized, allowing us to directly compare the effects of variables in terms of magnitude much more easily.

```{r}
#| echo: false
diabetes <- read.csv("diabetesdata.csv")
```

We begin by taking a brief look at the correlations in the data set. We observe that many of the blood serum measurements are moderately to strongly correlated, indicating the potential presence of multicollinearity.

```{r}
#| echo: false
#| label: fig-diab-cor
#| fig-cap: "Correlation plot of the Diabetes dataset"
#| message: false
#| fig-pos: "H"
library(corrplot)
cor_matrix <- cor(diabetes)
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45)
```

We begin by fitting an ordinary least squares regression using all predictors. The estimated model is summarized in @tbl-ols-mod, below. Among the non-blood serum measurements, all predictors except age are statistically significant. In contrast, within the blood serum measurements, only S5 shows significance.

```{r}
#| echo: false
#| label: tbl-ols-mod
#| tbl-cap: "OLS Regression Coefficients"
model <- lm(Y ~ ., data = diabetes)
coef_table <- summary(model)$coefficients
kable(coef_table, digits = 4)  # no caption inside kable


```

@tbl-ols-metrics, below, summarizes key metrics for the OLS model. With an $R^2$ around 0.5, the model explains a moderate proportion of the variation in disease progression for diabetes.

```{r}
#| echo: false
#| label: tbl-ols-metrics
#| tbl-cap: "OLS Model Summary Statistics"

# Extract key model statistics
model_stats <- data.frame(
  Statistic = c("Residual Standard Error", "Adjusted R-squared", "Multiple R-squared", "F-statistic"),
  Value = c(summary(model)$sigma,
            summary(model)$adj.r.squared,
            summary(model)$r.squared,
            summary(model)$fstatistic[1])
)

# Display model summary statistics as a kable
kable(model_stats, digits = 4)

```

We now move on to attempting to fit a model using regularized methods, beginning with Ridge regression.

```{r}
#| echo: false
set.seed(42)

# Load necessary libraries
library(ggplot2)
library(reshape2)

# Load data
diabetes <- read.csv("diabetesdata.csv")

# Separate predictors and response
X <- as.matrix(diabetes[, -which(names(diabetes) == "Y")])
y <- diabetes$Y


# Cross-validation to select lambda
set.seed(42)
cv_ridge <- cv.glmnet(X, y, alpha = 0, standardize = FALSE)

# Use the same lambda grid as in cv.glmnet to fit the model
ridge_model <- glmnet(X, y, alpha = 0, standardize = FALSE, lambda = cv_ridge$glmnet.fit$lambda)
```

The output below illustrates the Ridge Regression coefficients at $\lambda_{\min}$. We see that the blood serum measurements (along with sex, which is correlated with S3) have been shrunk to more similar levels compared to the OLS estimates, where a few coefficients were much larger.

```{r}
#| echo: false
coef(ridge_model, s = cv_ridge$lambda.min)
```

@fig-dcof-ridge, below, shows the coefficient paths for the ridge regression. We observe that the blood serum variables move together, while sex is shrunk quite aggressively.

```{r}
#| echo: false
#| label: fig-dcof-ridge
#| fig-cap: "Coefficient path for Ridge regression on Diabetes dataset."
#| warning: false
#| message: false
#| fig-pos: "H"

coef_mat <- as.matrix(ridge_model$beta)
coef_df <- data.frame(lambda = ridge_model$lambda, t(coef_mat))
coef_melt <- melt(coef_df, id.vars = "lambda")

ggplot(coef_melt, aes(x = -log(lambda), y = value, color = variable)) +
  geom_line(size = 1) +
  geom_vline(xintercept = -log(cv_ridge$lambda.min), linetype = "dashed", color = "red") +
  labs(x = "-log(lambda)", y = "Coefficient",
       title = "Ridge Regression Coefficient Paths",
       subtitle = "Red dashed line indicates lambda.min") +
  theme_minimal() +
  theme(legend.title = element_blank())

```

@fig-dcv-ridge, below, illustrates the cross-validation error path for the Ridge regression. We see that the error decreases smoothly, without a sharp plateau.

```{r}
#| echo: false
#| label: fig-dcv-ridge
#| fig-cap: "Cross validation error path for Ridge regression on Diabetes dataset."
#| fig-pos: "H"
# Prepare CV data
cv_df <- data.frame(
  lambda = cv_ridge$lambda,
  cvm = cv_ridge$cvm,   # mean CV error
  cvup = cv_ridge$cvup, # upper SE
  cvlo = cv_ridge$cvlo  # lower SE
)

# Plot CV curve with -log(lambda)
ggplot(cv_df, aes(x = -log(lambda), y = cvm)) +
  geom_line(size = 1, color = "blue") +
  geom_ribbon(aes(ymin = cvlo, ymax = cvup), alpha = 0.2) +
  geom_vline(xintercept = -log(cv_ridge$lambda.min), linetype = "dashed", color = "red") +
  labs(
    x = "-log(lambda)",
    y = "Mean CV Error",
    title = "Cross-Validation Curve for Ridge Regression",
    subtitle = "Red dashed = lambda.min"
  ) +
  theme_minimal()


```
We now move on to fitting the model using LASSO.
```{r}
#| echo: false
lasso_mod <- glmnet(X, y, alpha = 1, standardize = FALSE)


set.seed(42)
cv_lasso_mod <- cv.glmnet(X, y, alpha = 1, standardize = FALSE)

# Best lambda (lambda.min)
lambda_best <- cv_lasso_mod$lambda.min


coef_lasso <- coef(cv_lasso_mod, s = "lambda.min")

# Convert to data frame for reporting
coef_lasso_df <- data.frame(
  Variable = rownames(coef_lasso),
  Coefficient = as.numeric(coef_lasso)
)



```

The below output illustrates the estimated coefficients for the LASSO model. While some coefficients have been slightly shrunk, the estimates are overall quite similar to the OLS model, and no coefficients have been set to 0.

```{r}
#| echo: false
coef(lasso_mod, s = cv_lasso_mod$lambda.min)
```

@fig-dcof-lasso, below, illustrates the coefficient paths for LASSO. We see that at higher values of $\lambda$, many of the estimates are shrunk close to 0, but S5 and Sex quickly approach their OLS estimates once they enter the model.

```{r}
#| echo: false
#| label: fig-dcof-lasso
#| fig-cap: "Coefficient path for LASSO on Diabetes dataset."
#| fig-pos: "H"


# Extract coefficients from Lasso model
coef_mat <- as.matrix(lasso_mod$beta)
coef_df <- data.frame(lambda = lasso_mod$lambda, t(coef_mat))
coef_melt <- melt(coef_df, id.vars = "lambda")

# Plot with -log(lambda)
ggplot(coef_melt, aes(x = -log(lambda), y = value, color = variable)) +
  geom_line(size = 1) +
  geom_vline(xintercept = -log(cv_lasso_mod$lambda.min), linetype = "dashed", color = "red") +
  labs(x = "-log(lambda)", y = "Coefficient",
       title = "Lasso Regression Coefficient Paths",
       subtitle = "Red dashed line indicates lambda.min") +
  theme_minimal() +
  theme(legend.title = element_blank())


```

@fig-dcv-lasso, below, illustrates the cross-validation error path for LASSO. Unlike Ridge, the path has a large plateau. This suggests that using $\lambda_{1\text{se}}$ and evaluating the model on unseen data could be potentially beneficial, given how different the model appears at the start of the plateau compared to the end in the coefficient paths.

```{r}
#| echo: false
#| label: fig-dcv-lasso
#| fig-cap: "Cross validation error path for LASSO on Diabetes dataset."
#| fig-pos: "H"
# Prepare CV data
cv_df <- data.frame(
  lambda = cv_lasso_mod$lambda,
  cvm = cv_lasso_mod$cvm,   # mean CV error
  cvup = cv_lasso_mod$cvup, # upper SE
  cvlo = cv_lasso_mod$cvlo  # lower SE
)

# Plot CV curve with -log(lambda)
ggplot(cv_df, aes(x = -log(lambda), y = cvm)) +
  geom_line(size = 1, color = "blue") +
  geom_ribbon(aes(ymin = cvlo, ymax = cvup), alpha = 0.2) +
  geom_vline(xintercept = -log(cv_lasso_mod$lambda.min), linetype = "dashed", color = "red") +
  labs(
    x = "-log(lambda)",
    y = "Mean CV Error",
    title = "Cross-Validation Curve for Lasso Regression",
    subtitle = "Red dashed = lambda.min"
  ) +
  theme_minimal()

```

In order to evaluate the performance of the models, we performed 10-fold cross-validation (CV). In this procedure, the dataset is randomly split into 10 equally sized subsets, or “folds.” For each iteration, one fold is held out as a test set while the model is trained on the remaining nine folds. The trained model is then used to predict the responses for the held-out fold, and the mean squared error (MSE) is calculated. This process is repeated for all 10 folds, and the resulting MSE values are averaged to provide an estimate of the model’s predictive performance on unseen data.

For OLS, the model was directly fitted on the training folds in each iteration. For Ridge and Lasso, the cross-validation procedure used the previously selected optimal regularization parameter, $\lambda_{\min}$,⁡ obtained from CV on the full dataset. Since the best lambda was chosen using information from the entire dataset, this introduces a slight “data peeking”, meaning that the reported CV errors for Ridge and Lasso may be slightly optimistic.

For Lasso, it could also have been interesting to evaluate performance using $\lambda_{1\text{se}}$, which provides a more conservative choice that often results in sparser models. However, we did not use it in this analysis because it was not specified.

The resulting MSE values are described in @tbl-10fold-mse, below. We observe that OLS performed the best, with LASSO performing just slightly worse (which is natural, given that the shrinkage was very low). Ridge performs slightly worse than both.

```{r}
#| echo: false
#| label: tbl-10fold-mse
#| tbl-cap: "MSE from 10-fold CV"
#| message: false
#| warning: false
library(caret)

set.seed(42)

# 1. Prepare data
X <- as.matrix(diabetes[, -which(names(diabetes) == "Y")])
y <- diabetes$Y

# 2. Define 10-fold CV
folds <- createFolds(y, k = 10, list = TRUE, returnTrain = TRUE)

# -------------------------
# 3. OLS CV
# -------------------------
ols_mse_vec <- c()

for (i in 1:10) {
  train_idx <- folds[[i]]
  test_idx <- setdiff(1:nrow(X), train_idx)
  
  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  
  X_test <- X[test_idx, ]
  y_test <- y[test_idx]
  
  # Proper data frames for lm
  data_train <- data.frame(y = y_train, X_train)
  data_test  <- data.frame(y = y_test, X_test)
  
  ols_fit <- lm(y ~ ., data = data_train)
  y_pred <- predict(ols_fit, newdata = data_test)
  
  ols_mse_vec <- c(ols_mse_vec, mean((y_test - y_pred)^2))
}

ols_mse <- mean(ols_mse_vec)

# -------------------------
# 4. Ridge CV (alpha = 0)
# -------------------------
cv_ridge <- cv.glmnet(X, y, alpha = 0, standardize = FALSE)
ridge_best_lambda <- cv_ridge$lambda.min

ridge_mse_vec <- c()
for (i in 1:10) {
  train_idx <- folds[[i]]
  test_idx <- setdiff(1:nrow(X), train_idx)
  
  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  
  X_test <- X[test_idx, ]
  y_test <- y[test_idx]
  
  ridge_fit <- glmnet(X_train, y_train, alpha = 0, lambda = ridge_best_lambda, standardize = FALSE)
  y_pred <- predict(ridge_fit, newx = X_test)
  
  ridge_mse_vec <- c(ridge_mse_vec, mean((y_test - y_pred)^2))
}

ridge_mse <- mean(ridge_mse_vec)

# -------------------------
# 5. Lasso CV (alpha = 1)
# -------------------------
cv_lasso <- cv.glmnet(X, y, alpha = 1, standardize = FALSE)
lasso_best_lambda <- cv_lasso$lambda.min

lasso_mse_vec <- c()
for (i in 1:10) {
  train_idx <- folds[[i]]
  test_idx <- setdiff(1:nrow(X), train_idx)
  
  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  
  X_test <- X[test_idx, ]
  y_test <- y[test_idx]
  
  lasso_fit <- glmnet(X_train, y_train, alpha = 1, lambda = lasso_best_lambda, standardize = FALSE)
  y_pred <- predict(lasso_fit, newx = X_test)
  
  lasso_mse_vec <- c(lasso_mse_vec, mean((y_test - y_pred)^2))
}

lasso_mse <- mean(lasso_mse_vec)

# -------------------------
# 6. Combine results
# -------------------------
mse_comparison <- data.frame(
  Model = c("OLS", "Ridge", "Lasso"),
  CV_MSE = c(ols_mse, ridge_mse, lasso_mse)
)

kable(mse_comparison, digits = 2)


```

In terms of coefficient shrinkage, OLS does not apply any regularization. This can result in large differences in magnitude between predictors, particularly when some are correlated or have higher variance. Ridge regression, on the other hand, shrinks all coefficients towards zero, though none are set exactly to zero. In this dataset, the shrinkage is modest, but Ridge reduces extreme coefficient values, especially for correlated blood serum measurements (and *Sex*), providing more stable estimates. Lasso performs both shrinkage and variable selection, potentially setting some coefficients exactly to zero. However, at $\lambda_{\min}$ the shrinkage was small and no coefficients were removed, so the estimates remain quite similar to OLS. However, using a more conservative choice such as $\lambda_{1\text{se}}$ could perhaps produce a sparser, more interpretable model. On the other hand, eliminating correlated variables could make the model simpler, it would also discard useful information. Ideally, we would prefer to retain all informative predictors while mitigating the negative effects of multicollinearity through shrinkage or other methods.

Regarding model interpretability, OLS is straightforward, as all variables are retained and coefficients can be directly compared. Ridge is slightly less interpretable because coefficients are shrunk, but it does not perform variable selection, so all predictors remain in the model. Lasso is potentially the most interpretable, as it can highlight the most important predictors by eliminating uninformative ones, but in this analysis it was overall very similiar to OLS.

Given that the regularized methods provide only modest gains and no variables were actually eliminated, OLS might be considered the best choice for this dataset, producing unbiased estimates while retaining all informative predictors.

# Use of AI

AI (ChatGPT) has been utilized in this report primarily to assist with coding, including constructing and formatting plots, generating LaTeX code, and suggesting functions. It was also used to streamline the workflow, for example by automating the collection of estimates, parallelizing functions, and optimizing repetitive procedures. Additionally, it provided light spelling and grammar checks, as well as suggestions for more clearly conveying results.
